---
title: "High-level associations between measurement types and specific species / alpha-diversity"
output:
  html_document:
    df_print: paged
    code_folding: hide
---

In this notebook we explore whether certain bacterial species abundances were predictable using different types of available data. (For example: is there any specific species that can be well predicted given information about the dog's diet?). For this analysis only taxa with prevalence > 30% and a mean relative abundance > 0.05% were included. For each data type and each taxon, models were trained and evaluated using 5-fold cross validation, repeated 10 times. Spearman correlation was used to evaluate each model's performance on out-of-fold samples, comparing predicted relative abundances to true ones.

*** 

Preparations:

```{r message=FALSE, warning=FALSE}
# Preparations
library(tidyverse)
library(readr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
library(ape)
library(cowplot)
library(vegan)
library(ecodist)
library(stringr)
library(ranger)
library(metap)
library(rsample)
library(glmnet)
library(Matrix)
library(scales)

# Resolve possible conflicts
library(conflicted)
filter <- dplyr::filter
select <- dplyr::select

options(scipen = 999)
source('utils.R')

# Paths
bac120_taxonomy_r207_tsv <- "../GTDB/release207/bac120_taxonomy_r207.tsv"
ar53_taxonomy_r207_tsv <- "../GTDB/release207/ar53_taxonomy_r207.tsv"
dir.create("supp_tables")
```

Load data: 

```{r}
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

***

## Load data

Assumes that the "Data processing" section of the notebook `omic_and_data_type_associations.Rmd` was executed prior to running the below chunk.

```{r}
load('backup_data.RData')

# Apply a stronger filtering of rare features
prevalence_cutoff = 0.3
avg_abundance_cutoff = 0.0005

message('Removing rare phyla from Kraken tables')
ra_kraken_ph <- remove_rare_features(
  ra_kraken_ph, 
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff
  )
message('Removing rare genera from Kraken tables')
ra_kraken_ge <- remove_rare_features(
  ra_kraken_ge, 
  prevalence_cutoff=prevalence_cutoff, 
  avg_abundance_cutoff=avg_abundance_cutoff
  )
message('Removing rare species from Kraken tables')
ra_kraken_sp <- remove_rare_features(
  ra_kraken_sp, 
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff
  )
```

Compute average abundances of each taxon:

```{r}
ph_abunds <- ra_kraken_ph %>%
  tidyr::pivot_longer(cols = -dap_dog_id, names_to = 'ph', values_to = 'abun') %>%
  select(-dap_dog_id) %>%
  group_by(ph) %>%
  summarise(mean_abun = mean(abun), prev = sum(abun>0)/n(), n_samples = n())

ge_abunds <- ra_kraken_ge %>%
  tidyr::pivot_longer(cols = -dap_dog_id, names_to = 'ge', values_to = 'abun') %>%
  select(-dap_dog_id) %>%
  group_by(ge) %>%
  summarise(mean_abun = mean(abun), prev = sum(abun>0)/n(), n_samples = n())

sp_abunds <- ra_kraken_sp %>%
  tidyr::pivot_longer(cols = -dap_dog_id, names_to = 'sp', values_to = 'abun') %>%
  select(-dap_dog_id) %>%
  group_by(sp) %>%
  summarise(mean_abun = mean(abun), prev = sum(abun>0)/n(), n_samples = n())
```

Measurement types used in this analysis:

```{r}
# Pretty names for plots
pretty_names <- c("main_demographics" = 
                  "Age, Weight,\nSex, Breed",
                  "lab_test" = "Blood tests",
                  "environment_data" = "Environment", 
                  "pa_data" = "Lifestyle",
                  "diet_data" = "Diet",
                  "clinical_data" = "Health\nconditions",
                  "sample_qc_data" = "Seq' quality")

all_meas_types <- names(pretty_names)
```

Load a GTDB mapping file (to map each clade to higher-level taxonomic groups):

```{r}
bac120_taxonomy_r207 <- read_delim(
  bac120_taxonomy_r207_tsv, 
  delim = ";", 
  escape_double = FALSE, 
  col_names = c('ki','ph','cl','or','fa','ge','sp'), 
  trim_ws = TRUE,
  show_col_types = FALSE
  ) %>% mutate(ki = 'Bacteria')

ar53_taxonomy_r207 <- read_delim(
  ar53_taxonomy_r207_tsv, 
  delim = ";", 
  escape_double = FALSE, 
  col_names = c('ki','ph','cl','or','fa','ge','sp'), 
  trim_ws = TRUE,
  show_col_types = FALSE
  ) %>% mutate(ki = 'Archaea')
  
gtdb_tax <- bind_rows(bac120_taxonomy_r207, ar53_taxonomy_r207) %>%
  mutate(sp = gsub('^s__','',sp)) %>%
  mutate(ge = gsub('^g__','',ge)) %>%
  mutate(fa = gsub('^f__','',fa)) %>%
  mutate(or = gsub('^o__','',or)) %>%
  mutate(cl = gsub('^c__','',cl)) %>%
  mutate(ph = gsub('^p__','',ph))

tmp_fun <- function(x) { gsub('\\-| ', '\\.', x) }
gtdb_tax <- gtdb_tax %>% mutate_all(tmp_fun)

ge_to_ph <- gtdb_tax %>%
  select(ge, ph) %>%
  distinct()

ge_abunds <- ge_abunds %>% left_join(ge_to_ph, by = 'ge')
```

***

# Alpha-diversity predictability

```{r}
alpha <- map_sample_id_to_dap_dog_id(shannon, sample_map)
```


```{r}
backup_file <- 'omic_associations_backup/alpha_preds_backup.csv'
if (file.exists(backup_file)) {
  message('Loading alpha-diversity modelling results from backup')
  alpha_preds <- read_csv(backup_file, show_col_types = FALSE)
} else {
  # Initialize empty table
  alpha_preds <- data.frame()
  alpha_var_imp <- data.frame()
  
  # Iterate over each measurement type
  for (curr_features in all_meas_types) {
    message("Working on measurement-type: ", curr_features)
    curr_features2 <- get(curr_features) %>% na.omit()
    
    # Train & evaluate RF models
    tmp_res <- get_predictability_regression(
      feat_table = curr_features2, 
      metadata = alpha, 
      sample_id_column = 'dap_dog_id', 
      vars_to_analyze = c("shannon_sp"),
      ml_method = 'rf',
      corr_method = "spearman",
      corr_threshold = 0.2,
      n_ml_repeats = 10,
      n_ml_folds = 5,
      feat_importance = 'none',
      rand_seed = 27
      )
    
    # Record all results
    alpha_preds <- bind_rows(
      alpha_preds,
      tmp_res$results %>% 
        mutate(measurement_type = pretty_names[curr_features]) %>%
        mutate(n_features_in_model = ncol(curr_features2)-1) %>%
        select(-Significant)
    )
    message('----------------------------------')
  } 
  
  # Re-calculate FDR
  alpha_preds$new_fdr <- p.adjust(alpha_preds$p, method = 'fdr')
  
  # Save to avoid re-computing each time
  write_csv(alpha_preds, file = backup_file)
}
```

```{r fig.width=5, fig.height=4}
get_signif_mark <- function(x) {if(x<0.001) return('***'); if(x<0.01) return('**'); if(x<0.05) return('*'); if(x<0.1) return('.'); return('')}

tmp <- alpha_preds %>%
  filter(measure == 'shannon_sp') %>%
  arrange(mean_cor)
tmp$label <- sapply(tmp$new_fdr, FUN = get_signif_mark)
tmp$measurement_type <- factor(tmp$measurement_type, levels = rev(tmp$measurement_type))

ggplot(tmp, aes(x = measurement_type, y = mean_cor, fill = mean_cor, label = label)) +
  geom_col(color = 'black', width = 0.6) +
  geom_text(vjust = 0) +
  scale_fill_gradient(name = 'Correlation r', low = 'grey90', high = 'orchid4') +
  scale_y_continuous(expand=expansion(mult=c(0,0.1))) +
  geom_hline(yintercept = 0, linetype = 'dashed') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  xlab(NULL) +
  ylab('Model performance\n(Correlation between observed\nand predicted alpha diversity)') +
  ggtitle('Potential determinants of shannon diversity') +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = 'none')

```

***

# Genera predictability 

```{r}
# Take 20 top abundant genera only
top_abundant_genera <- ge_abunds %>%
  top_n(n = 20, wt = mean_abun) %>%
  arrange(-mean_abun)
```


```{r}
backup_file <- 'omic_associations_backup/all_tax_preds_backup.csv'
if (file.exists(backup_file)) {
  message('Loading modelling results from backup')
  all_tax_preds <- read_csv(backup_file, show_col_types = FALSE)
} else {
  # Initialize empty table
  all_tax_preds <- data.frame()
  
  # Iterate over each measurement type
  for (curr_features in all_meas_types) {
    message("Working on measurement-type: ", curr_features)
    
    # Pull data
    curr_features_df <- get(curr_features) %>% na.omit()
    
    # Train & evaluate RF models
    tmp_res <- get_predictability_regression(
      feat_table = curr_features_df, 
      metadata = ra_kraken_ge, 
      sample_id_column = 'dap_dog_id', 
      vars_to_analyze = top_abundant_genera$ge,
      ml_method = 'rf',
      corr_method = "spearman",
      corr_threshold = 0.2,
      n_ml_repeats = 10,
      n_ml_folds = 5,
      rand_seed = 27
      )
    
    # Record all results
    all_tax_preds <- bind_rows(
      all_tax_preds,
      tmp_res$results %>% 
        mutate(taxa_level = 'ge') %>%
        mutate(measurement_type = pretty_names[curr_features]) %>%
        mutate(n_features_in_model = ncol(curr_features_df)-1) %>%
        select(-Significant)
    )
    message('----------------------------------')
  } 
  
  # Re-calculate FDR
  all_tax_preds$new_fdr <- p.adjust(all_tax_preds$p, method = 'fdr')
  
  # Add phyla
  all_tax_preds <- all_tax_preds %>% left_join(ge_to_ph, by = c('measure'='ge'))
  
  # Save to avoid re-computing each time
  write_csv(all_tax_preds, file = backup_file)
}
```
      
## Plot

```{r fig.width=7, fig.height=4.5}
fdr_threshold_grey_out <- 0.05
cor_thres_grey_out <- 0

# Merge both genera predictions and shannon diversity predictions
tmp1 <- all_tax_preds %>%
  filter(taxa_level == 'ge') %>%
  select(measure, measurement_type, new_fdr, mean_cor, ph) 

tmp2 <- alpha_preds %>%
  filter(measure == 'shannon_sp') %>%
  mutate(measure = 'Shannon diversity') %>%
  select(measure, measurement_type, new_fdr, mean_cor) %>%
  mutate(ph = '')

tmp <- bind_rows(tmp1, tmp2)
# Add significant marks
tmp$label <- sapply(tmp$new_fdr, FUN = get_signif_mark)

# Define abbreviations for phyla names
ph_abbrev <- c('Actinobacteriota'='Ac.',
               'Campylobacterota'='Cam.',
               'Firmicutes_C'='Fir.C',
               'Fusobacteriota'='Fus.',
               #'Firmicutes_A'='Firm._A',
               'Firmicutes'='Fir.',
               'Proteobacteria'='Pro.')

# Plot
p_abunds <- ggplot(
  ge_abunds %>%
    filter(ge %in% top_abundant_genera$ge) %>%
    # Use abbreviations
    mutate(ph = ifelse(ph %in% names(ph_abbrev), ph_abbrev[ph], ph)) %>%
    # Just in order to align correctly to the graph below
    bind_rows(data.frame(ge = 'Shannon diversity', ph = '', mean_abun = 0)), 
                   aes(x = ge, y = mean_abun)) +
  geom_col(color = 'black', width = 0.8) +
  ylab('Mean relative\nabundance') +
  xlab(NULL) +
  theme_minimal() +
  facet_grid(~ph, scales = 'free', space = 'free') +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x = element_blank()) +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme(plot.margin = unit(c(5, 5, 0, 5), "points")) +
  theme(strip.text = element_blank())

p_grid <- ggplot(
  tmp %>%
    # Only top abundant species
    filter(measure %in% top_abundant_genera$ge | measure == 'Shannon diversity') %>%
    # Use abbreviations
    mutate(ph = ifelse(ph %in% names(ph_abbrev), ph_abbrev[ph], ph)) %>%
    # Grey out FDR's above a threshold
    mutate(mean_cor = ifelse(
      new_fdr <= fdr_threshold_grey_out & 
        mean_cor >= cor_thres_grey_out, 
      mean_cor, NA
      )), 
    aes(x = measure, 
        y = measurement_type, 
        fill = mean_cor,
        label = label
        )) +
  geom_tile(linewidth = 1.1, linejoin = 'mitre', color = 'white') +
  geom_text(size = 3) +
  scale_fill_gradientn(
    name = expression(rho),  #'Rho',
    na.value = 'grey90',
    colors = c('white','#F5E8B7','#CD5C08','darkred'),
    limits = c(cor_thres_grey_out - 0.01, NA),
    breaks = c(0,0.05,0.1,0.15,0.2,0.25,0.3,0.35),
    guide = guide_colorbar(
      frame.colour = "black", 
      ticks.colour = "black"
      )) +
  scale_alpha_manual(values = 1, labels = 
                       paste('FDR >', fdr_threshold_grey_out)) +
  xlab(NULL) +
  ylab(NULL) +
  theme_minimal() +
  facet_grid(~ph, scales = 'free', space = 'free', switch = 'x') +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.margin = margin(-0.3,0,0,0, unit="cm")) +
  theme(plot.margin = unit(c(0, 5, 0, 5), "points")) +
  theme(strip.placement = "outside") +
  theme(strip.text = element_text(face='bold', color='darkred')) +
  theme(legend.key.width= unit(0.3, 'cm'))

p <- plot_grid(p_abunds, p_grid, 
          ncol = 1, 
          rel_heights = c(2,5), 
          align = 'v', axis = 'lr')
p
```

```{r}
ggsave(file="Figure2/Figure2D.svg", plot=p, width=7.5, height=4.5)
```


## Supplementary Tables

### S1: Average taxa abundances

Note: only taxa with a prevalence over 30% are included here

```{r}
bind_rows(
  ph_abunds %>% rename(Taxon = ph) %>% mutate(tax_level = 'Phylum'),
  ge_abunds %>% rename(Taxon = ge) %>% mutate(tax_level = 'Genus') %>% select(-ph),
  sp_abunds %>% rename(Taxon = sp) %>% mutate(tax_level = 'Species'),
) %>%
  mutate(mean_abun = round(mean_abun, 5)) %>%
  mutate(prev = round(prev, 5)) %>%
  relocate(tax_level, Taxon, n_samples) %>%
  rename(`Taxonomic rank` = tax_level) %>%
  rename(`Average abundance` = mean_abun) %>%
  rename(`Prevalence` = prev) %>%
  rename(`Total number of samples analyzed` = n_samples) %>%
  write_csv('supp_tables/s1_taxa_average_abundances.csv')
```

### S5: Genera + alpha-diversity predictions based on different measurment types

```{r}
tmp %>%
  mutate(mean_cor = round(mean_cor, 5)) %>%
  mutate(new_fdr = round(new_fdr, 8)) %>%
  mutate(measurement_type = gsub('\n',' ',measurement_type)) %>%
  select(measure, measurement_type, mean_cor, new_fdr) %>%
  rename(`Feature predicted (genus abundance / alpha-diversity)` = measure) %>%
  rename(`Data type used as predictors` = measurement_type) %>%
  rename(`Spearman correlation - predicted vs. real (mean over folds)` = mean_cor) %>%
  rename(FDR = new_fdr) %>%
  write_csv('supp_tables/s5_genera_predictions.csv')
```



         