---
title: "DAP dogs' Precision 1 cohort description"
author: "Borenstein Lab"
date: "April 2025"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
---

In this file we describe the dogs in the DAP Precision 1 cohort and generate Figure 1.
The data is from the Terra release 2023_12022024.


```{r echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

# Import libraries

```{r message=FALSE, warning=FALSE}
# Preparations
library(readr)
library(ggplot2)
library(ggridges)
library(gridExtra)
library(tibble)
library(ggpubr)
library(tidyr)
library(dplyr)
library(themis)
library(ape)
library(cowplot)
library(vegan)
library(kableExtra)
library(stringr)
library(recipes)
library(ROSE)
library(forcats)
library(combinat)
library(lightgbm)
library(caret)
library(ranger)
library(randomForest)
library(tidymodels)
options(scipen = 999)
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

# Load data

```{r}
dog_and_owner <- read_csv(file.path(metadata_dir, "terra_release_2023_12022024", "DAP_2023_HLES_dog_owner_v1.0.csv"),
                         show_col_types = FALSE,
                         guess_max = 50000) %>%
  rename(dap_dog_id = dog_id) %>%
  mutate(dap_dog_id = as.character(dap_dog_id))

# Merge the sample_,ap and the HLES datça
sample_map <- sample_map %>%
  inner_join(dog_and_owner, by="dap_dog_id")
```

Plot the number of dogs per state

```{r}

library(usmap)
states_map <- us_map()

country_counts <- sample_map %>%
  group_by(dd_acquired_state) %>%
  summarise(count = n())

map_data <- states_map %>%
  left_join(country_counts, by = c("abbr" = "dd_acquired_state"))

p <- ggplot(map_data) +
  geom_sf(aes(fill = count)) +
  scale_fill_viridis_c(direction = -1, option = "F", begin=0.3, name="Dogs", na.value = "white") +
  theme_classic() +
  theme(
    axis.title = element_blank(),       # Remove axis titles
    axis.text = element_blank(),        # Remove axis text
    axis.ticks = element_blank(),       # Remove axis ticks
    axis.line = element_blank(),        # Remove axis lines
    panel.grid = element_blank(),       # Remove grid lines (optional)
    axis.title.y = element_blank()      # Specifically remove y-axis title (if needed)
  )

p # Figure 1A
```


Plot the age distribution by age

```{r, fig.width=5, fig.height=3}
p <- sample_map %>%
  mutate(Sex = case_when(
    grepl("Female", Sex_class, ignore.case = TRUE) ~ "Female",
    grepl("Male", Sex_class, ignore.case = TRUE) ~ "Male"
  )) %>%
  ggplot(aes(x = age_at_stool_collection, fill=Sex)) +
  geom_density(alpha=0.5) +
  theme_classic() +
  xlab("Age (at stool collection)") +
  ylab("Density") +
  scale_fill_manual(values = c("#da9f93", "#7ca982")) +
  theme(
    legend.position = c(0.7, 1),            # Position legend inside plot, upper-left
    legend.justification = c(0, 1),           # Anchor legend by upper-left corner
    legend.background = element_rect(fill = "white", color = NA),  # Optional: white background for better visibility
    axis.text = element_text(size = 14),    # Enlarge axis text
    axis.title = element_text(size = 16),   # Enlarge axis titles
    legend.text = element_text(size = 12),  # Enlarge legend text
    legend.title = element_text(size = 14)  # Enlarge legend title
  )

p # Figure 1B
```

Plot the sex distribution (including intact/sterilized)

```{r, fig.width=6, fig.height=3.5}
p <- sample_map %>%
  mutate(Sex = case_when(
    grepl("Female", Sex_class, ignore.case = TRUE) ~ "Female",
    grepl("Male", Sex_class, ignore.case = TRUE) ~ "Male"
  )) %>%
  mutate(Sterilized = case_when(
    grepl("intact", Sex_class, ignore.case = TRUE) ~ "intact",
    TRUE ~ "sterilized"
  )) %>%
  ggplot(aes(x = Sex, fill=Sex_class)) +
  scale_fill_manual(values = c("#ebd4cb", "#da9f93", "#e0eec6", "#7ca982"), name="Sterilization status") +
  geom_bar(alpha=1, color="black") +
  xlab("Sex") +
  ylab("Dogs") +
  theme_classic()  +
  theme(
    axis.text = element_text(size = 14),    # Enlarge axis text
    axis.title = element_text(size = 16),   # Enlarge axis titles
    legend.text = element_text(size = 12),  # Enlarge legend text
    legend.title = element_text(size = 14)  # Enlarge legend title
  )

p # Figure 1C
```

Plot the breed distribution of the top 10 most common breeds

```{r, fig.width=4, fig.height=4}


breed_data <- read_delim(
  file.path(metadata_dir, terra_release, "DogAgingProject_GeneticData_CuratedRelease_2023.ancestry.tsv"),
  delim = "\t", 
  escape_double = FALSE, 
  trim_ws = TRUE,
  show_col_types = FALSE
  ) %>%
  mutate(dog = as.character(dog)) %>%
  mutate(pop = make.names(pop)) %>%
  select(-code) %>%
  # Remove rare breeds
  group_by(pop) %>%
  filter(n() >= 50) %>%
  ungroup() %>%
  filter(pct > 0.6) %>%
  inner_join(sample_map, by = c("dog" = "dap_dog_id")) 


# Filter for the top 10 most common categories and reverse the order
top_10_breed_data <- breed_data %>%
  mutate(pop = gsub("\\.", " ", pop)) %>%
  mutate(pop = fct_infreq(pop)) %>%
  filter(pop %in% levels(pop)[1:10]) %>%
  mutate(pop = fct_rev(pop))  # Reverse the order of the factor levels
         

# Plot the filtered data with the plot rotated
p <- top_10_breed_data %>%
  ggplot(aes(x = pop)) +
  geom_bar(fill = "#eddcd2", color="black", alpha=1) +
  coord_flip() +  # Rotate the entire plot
  labs(x = "Breed") +
  labs(y = "Dogs") +
  theme(axis.text.y = element_text(angle = 0, hjust = 1)) + 
  theme_classic() +
  theme(
    axis.text = element_text(size = 14),    # Enlarge axis text
    axis.title = element_text(size = 16),   # Enlarge axis titles
    legend.text = element_text(size = 12),  # Enlarge legend text
    legend.title = element_text(size = 14)  # Enlarge legend title
  )


p # Figure 1D
```

Plot the general health distribution by life stage

```{r, fig.width=4, fig.height=3.5}

p <- sample_map %>%
  drop_na(life_stage_at_stool_collection, hs_general_health) %>%
  mutate(hs_general_health = as.character(hs_general_health)) %>%
  mutate(hs_general_health = recode(hs_general_health,
                             "1" = "Excellent",
                             "2" = "Very good",
                             "3" = "Good",
                             "4" = "Fair",
                             "5" = "Poor",
                             .default = hs_general_health)) %>% # Retain other values unchanged
  mutate(life_stage_at_stool_collection = factor(life_stage_at_stool_collection, levels = c("Puppy", "Young Adult", "Mature Adult", "Senior")),
         hs_general_health = factor(hs_general_health, levels = c("Excellent", "Very good", "Good", "Fair", "Poor"))) %>%
  ggplot(aes(x=hs_general_health, group=life_stage_at_stool_collection, fill=life_stage_at_stool_collection)) +
  scale_fill_manual(values = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"), name="Life stage") +
  geom_bar(color="black", alpha=1) +
  xlab("General health") +
  ylab("Dogs") +
  theme_classic() +
  theme(
    legend.position = c(0.55, 1),            # Position legend inside plot, upper-left
    legend.justification = c(0, 1),           # Anchor legend by upper-left corner
    legend.background = element_rect(fill = "white", color = NA),  # Optional: white background for better visibility
    axis.text = element_text(size = 14),    # Enlarge axis text
    axis.title = element_text(size = 16),   # Enlarge axis titles
    legend.text = element_text(size = 12),  # Enlarge legend text
    legend.title = element_text(size = 14)  # Enlarge legend title
  )


p # Figure 1E

```

Plot the breed size distribution by life stage

```{r, fig.width=6, fig.height=3.5}
p <- sample_map %>%
  drop_na(life_stage_at_stool_collection) %>%
  mutate(life_stage_at_stool_collection = factor(life_stage_at_stool_collection, levels = c("Puppy", "Young Adult", "Mature Adult", "Senior")),
         breed_size = factor(breed_size, levels = c("Toy and Small", "Standard", "Medium", "Large", "Giant"))) %>%
  ggplot(aes(x = life_stage_at_stool_collection, fill=breed_size)) +
  scale_fill_manual(values = c("#c8553d", "#f28f3b", "#ffd5c2", "#ecf8f8", "#588b8b"), name="Breed size") +
  geom_bar(color="black", alpha=1) +
  # coord_flip() +  # Rotate the entire plot
  xlab("Life stage (at stool collection)") +
  ylab("Dogs") +
  theme_classic() +
  theme(
    legend.position = c(0.03, 1),            # Position legend inside plot, upper-left
    legend.justification = c(0, 1),         # Anchor legend by upper-left corner
    legend.background = element_rect(fill = "white", color = NA),  # Optional: white background for better visibility
    axis.text = element_text(size = 14),    # Enlarge axis text
    axis.title = element_text(size = 16),   # Enlarge axis titles
    legend.text = element_text(size = 12),  # Enlarge legend text
    legend.title = element_text(size = 14)  # Enlarge legend title
  ) 


p # Figure 1F
```
