---
title: "Comparison between DAP and other dog cohorts"
output:
  html_document:
    df_print: paged
---

In this notebook we compared DAP microbiome profiles to those from a previously published cohort - [Coehlo et al. 2018](https://microbiomejournal.biomedcentral.com/articles/10.1186/s40168-018-0450-3). Coehlo et al. raw metagenomic samples were re-processed using the same pipeline used for DAP samples. The code below produces Figures 2A, 2B and Supplementary Figure S1 in the manuscript.

Preparations:

```{r warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(tibble)
library(vegan)
library(ape)
library(cowplot)
library(scales)
options(scipen = 999)

# Location of Coehlo tables
CO_FOLDER <- "../data/coehlo_et_al_2018"
CO_METADATA <- file.path(CO_FOLDER, "table_s1.csv")
CO_READS <- file.path(CO_FOLDER, "table_s2.csv")
CO_SPECIES <- file.path(CO_FOLDER, "feature_tables", "kraken_species_level_taxonomy_RA.tsv")
CO_GENERA <- file.path(CO_FOLDER, "feature_tables", "kraken_genus_level_taxonomy_RA.tsv")
CO_PHYLA <- file.path(CO_FOLDER, "feature_tables", "kraken_phylum_level_taxonomy_RA.tsv")

dir.create("SuppFigure1")
dir.create("Figure2")
```

Load utility functions:

```{r}
source('utils.R')
```

# Load data 

## Load DAP profiles

```{r DAP_DATA_LOADING, warning=FALSE, message=FALSE, cache = TRUE}
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

## Load Coehlo metadata

```{r, cache = TRUE}
coe_metadata <- read_csv(CO_METADATA, show_col_types = F)
reads <- read_csv(CO_READS, show_col_types = F)

coe_metadata <- coe_metadata %>%
  filter(Diet == 'Baseline') %>%
  left_join(reads, by = "sample_id") %>%
  select(-sample_id, -Diet, -`Final fecal date`, -Note) %>%
  rename(sample_id = ENA_ACCESSION) %>%
  relocate(sample_id)

rm(reads)
```

Unify common metadata fields:

```{r}
metadata_all <- bind_rows(
  coe_metadata %>%
    select(sample_id, Age, Breed, Sex, Phenotype) %>%
    mutate(Cohort = 'COE') %>%
    mutate(Breed = ifelse(Breed == 'LAB', 'Labrador Retriever', Breed)) %>%
    mutate(Breed = ifelse(Breed == 'BEA', 'Beagle', Breed)),
  sample_map %>%
    mutate(Sex = ifelse(grepl('Female',Sex_class), 'F', 'M')) %>%
    select(dap_dog_id, age_at_stool_collection, Breed, Sex) %>%
    mutate(Cohort = 'DAP') %>%
    rename(sample_id = dap_dog_id, Age = age_at_stool_collection)
)
```

LAB = Labrador retrievers 
BEA = Beagles

```{r}
# Print statistics: number of counts from each cohort
table(metadata_all$Cohort)
```

## Load rarefied bracken tables (sp + ge + ph)

### DAP

```{r}
message("Mapping sample ID's to dog ID's")
ra_kraken_sp_dap <- map_sample_id_to_dap_dog_id(ra_kraken_sp, sample_map)
ra_kraken_ge_dap <- map_sample_id_to_dap_dog_id(ra_kraken_ge, sample_map)
ra_kraken_ph_dap <- map_sample_id_to_dap_dog_id(ra_kraken_ph, sample_map)
rm(ra_kraken_sp, ra_kraken_ph, ra_kraken_ge)
```

### Coehlo

```{r, cache = TRUE}
kraken_sp_coe <- read_tsv2(CO_SPECIES) %>% reorganize_kraken_table()
kraken_ge_coe <- read_tsv2(CO_GENERA) %>% reorganize_kraken_table()
kraken_ph_coe <- read_tsv2(CO_PHYLA) %>% reorganize_kraken_table()
```

```{r}
message('Converting Kraken tables to relative abundances')
ra_kraken_ph_coe <- convert_to_relab(kraken_ph_coe)
ra_kraken_ge_coe <- convert_to_relab(kraken_ge_coe)
ra_kraken_sp_coe <- convert_to_relab(kraken_sp_coe)
```

```{r, cache = TRUE}
prevalence_cutoff = 0.2
avg_abundance_cutoff = 0.0001

message('Removing rare phyla from Kraken tables')
ra_kraken_ph_coe <- remove_rare_features(
  ra_kraken_ph_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Removing rare genera from Kraken tables')
ra_kraken_ge_coe <- remove_rare_features(
  ra_kraken_ge_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Removing rare species from Kraken tables')
ra_kraken_sp_coe <- remove_rare_features(
  ra_kraken_sp_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Re-normalizing to relative abundances')
ra_kraken_ph_coe <- convert_to_relab(ra_kraken_ph_coe)
ra_kraken_ge_coe <- convert_to_relab(ra_kraken_ge_coe)
ra_kraken_sp_coe <- convert_to_relab(ra_kraken_sp_coe)
```

## Merge DAP + Coehlo

```{r, cache = TRUE}
all_ph <- bind_rows(
  ra_kraken_ph_coe %>%
    mutate(sample_id = paste('COE', sample_id, sep = '_')),
  ra_kraken_ph_dap %>%
    mutate(sample_id = paste('DAP', dap_dog_id, sep = '_')) %>%
    select(-dap_dog_id)
) 
all_ph[is.na(all_ph)] <- 0 

ph_in_coe <- names(ra_kraken_ph_coe)
ph_in_dap <- names(ra_kraken_ph_dap)
message(length(intersect(ph_in_coe, ph_in_dap)), ' of ',
        length(ph_in_coe), 
        ' phyla identified in Coehlo data are also in DAP (',
        round(100*length(intersect(ph_in_coe, ph_in_dap)) /
                length(ph_in_coe),2), ' %)')
message(length(intersect(ph_in_coe, ph_in_dap)), ' of ',
        length(ph_in_dap), 
        ' phyla identified in DAP data are also in Coehlo (',
        round(100*length(intersect(ph_in_dap, ph_in_coe)) /
                length(ph_in_dap),2), ' %)')
```

```{r, cache = TRUE}
all_ge <- bind_rows(
  ra_kraken_ge_coe %>%
    mutate(sample_id = paste('COE', sample_id, sep = '_')),
  ra_kraken_ge_dap %>%
    mutate(sample_id = paste('DAP', dap_dog_id, sep = '_')) %>%
    select(-dap_dog_id)
) 
all_ge[is.na(all_ge)] <- 0 

ge_in_coe <- names(ra_kraken_ge_coe)
ge_in_dap <- names(ra_kraken_ge_dap)
message(length(intersect(ge_in_coe, ge_in_dap)), ' of ',
        length(ge_in_coe), 
        ' non-rare genera identified in Coehlo data are also in DAP (',
        round(100*length(intersect(ge_in_coe, ge_in_dap)) /
                length(ge_in_coe),2), ' %)')
message(length(intersect(ge_in_coe, ge_in_dap)), ' of ',
        length(ge_in_dap), 
        ' non-rare genera identified in DAP data are also in Coehlo (',
        round(100*length(intersect(ge_in_coe, ge_in_dap)) /
                length(ge_in_dap),2), ' %)')
```

```{r, cache = TRUE}
all_sp <- bind_rows(
  ra_kraken_sp_coe %>%
    mutate(sample_id = paste('COE', sample_id, sep = '_')),
  ra_kraken_sp_dap %>%
    mutate(sample_id = paste('DAP', dap_dog_id, sep = '_')) %>%
    select(-dap_dog_id)
) 
all_sp[is.na(all_sp)] <- 0 

sp_in_coe <- names(ra_kraken_sp_coe)
sp_in_dap <- names(ra_kraken_sp_dap)
message(length(intersect(sp_in_coe, sp_in_dap)), ' of ',
        length(sp_in_coe), 
        ' non-rare species identified in Coehlo data are also in DAP (',
        round(100*length(intersect(sp_in_coe, sp_in_dap)) /
                length(sp_in_coe),2), ' %)')
message(length(intersect(sp_in_coe, sp_in_dap)), ' of ',
        length(sp_in_dap), 
        ' non-rare species identified in DAP data are also in Coehlo (',
        round(100*length(intersect(sp_in_coe, sp_in_dap)) /
                length(sp_in_dap),2), ' %)')
```

## Counts of shared taxa

```{r fig.width=4.5, fig.height=1.5}
# Plot taxa overlap
tmp <- bind_rows(
  data.frame(
    tax_level = 'Species',
    category = c('DAP only','Shared','Coehlo only'),
    count = c(sum(! sp_in_dap %in% sp_in_coe),
              sum(sp_in_dap %in% sp_in_coe),
              sum(! sp_in_coe %in% sp_in_dap))
  ),
  data.frame(
    tax_level = 'Genera',
    category = c('DAP only','Shared','Coehlo only'),
    count = c(sum(! ge_in_dap %in% ge_in_coe),
              sum(ge_in_dap %in% ge_in_coe),
              sum(! ge_in_coe %in% ge_in_dap))
  ),
  data.frame(
    tax_level = 'Phyla',
    category = c('DAP only','Shared','Coehlo only'),
    count = c(sum(! ph_in_dap %in% ph_in_coe),
              sum(ph_in_dap %in% ph_in_coe),
              sum(! ph_in_coe %in% ph_in_dap))
  )
) %>%
  mutate(category = factor(category, levels = c('DAP only','Shared','Coehlo only'))) %>%
  mutate(tax_level = factor(
    tax_level,
    levels = c('Phyla','Genera','Species'),
    labels = paste0(c('Phyla','Genera','Species'),' [N=',c(length(union(ph_in_coe,ph_in_dap)),
                                                          length(union(ge_in_coe,ge_in_dap)),
                                                          length(union(sp_in_coe,sp_in_dap))),']'))) %>%
  group_by(tax_level) %>%
  mutate(perc = 100*count/sum(count)) %>%
  ungroup() %>%
  mutate(perc_label = paste0(round(perc,1),'%')) 

p <- ggplot(tmp, aes(x=tax_level, y=perc/100, fill = category, alpha = category)) +
  geom_bar(stat = 'identity', color = 'black', width = 0.8) +
  theme_classic() + 
  coord_flip() +
  xlab(NULL) +
  ylab('Percent of (non-rare) taxa') +
  scale_fill_manual('', values = c('deepskyblue4', '#3E1842', 'darkred')) +
  scale_alpha_manual('', values = c(0.5,0.8,0.7)) +
  scale_y_continuous(expand = c(0,0), labels = scales::percent) +
  guides(fill = guide_legend(reverse=TRUE), alpha = guide_legend(reverse=TRUE)) +
  theme(legend.position = 'top') +
  theme(legend.key.size = unit(12,"points")) +
  theme(legend.box.margin=margin(-10,-10,-10,-10)) +
  theme(plot.margin = unit(c(5, 15, 5, 5), "points")) +
  theme(panel.grid.major.x = element_line(color = 'grey80'))

p
ggsave(file="SuppFigure1/SuppFigure1B.svg", plot=p, width=4.5, height=1.5)
```

# Compute Bray-Curtis (sp- and ge-level)

We compute two versions: (1) with all taxa, even if they appear in only a single cohort, and (2) only with shared taxa.

```{r, cache = TRUE}
bc_ge <- vegdist(all_ge %>% column_to_rownames('sample_id')) # Method = BC
bc_sp <- vegdist(all_sp %>% column_to_rownames('sample_id')) 

bc_ge_common <- vegdist(all_ge %>% column_to_rownames('sample_id') %>% 
                          select(all_of(intersect(ge_in_coe, ge_in_dap)))) 
bc_sp_common <- vegdist(all_sp %>% column_to_rownames('sample_id') %>% 
                          select(all_of(intersect(sp_in_coe, sp_in_dap)))) 

# Reorder samples so that Coehlo samples appear on top
bc_sp <- as.dist(as.matrix(bc_sp)[rev(labels(bc_sp)),rev(labels(bc_sp))])
bc_ge <- as.dist(as.matrix(bc_ge)[rev(labels(bc_ge)),rev(labels(bc_ge))])
bc_sp_common <- as.dist(as.matrix(bc_sp_common)[rev(labels(bc_sp_common)),rev(labels(bc_sp_common))])
bc_ge_common <- as.dist(as.matrix(bc_ge_common)[rev(labels(bc_ge_common)),rev(labels(bc_ge_common))])
```


# PCoA

## Species level

```{r}
plot_pcoa_coe_vs_dap <- function(pcs, metadata_all) {
  pcs_to_plot <- pcs$vectors %>% 
    as.data.frame() %>%
    select(Axis.1, Axis.2, Axis.3, Axis.4) %>%
    rownames_to_column('sample_id_tmp') %>%
    tidyr::separate(col = sample_id_tmp, 
                    into = c('Cohort','sample_id')) %>%
    left_join(metadata_all, by = c('Cohort','sample_id'))
  
  label_x <- paste0('1st PC - ', 
                    round(pcs$values$Rel_corr_eig[1]*100,1), '% explained variance')
  label_y <- paste0('2nd PC - ', 
                    round(pcs$values$Rel_corr_eig[2]*100,1), '% explained variance')
  
  p_pcoa <- ggplot(pcs_to_plot, 
           aes(x = Axis.1, y = Axis.2, 
               color = Cohort)) +
    geom_point(alpha = 0.3, size = 2) +
    xlab(label_x) +
    ylab(label_y) +
    scale_color_manual(
      values = c('COE' = 'darkred','DAP' = 'deepskyblue4'),
      guide = guide_legend(keywidth = 0.1, override.aes = list(alpha = 1, size = 3))
      ) +
    theme_classic() + 
    theme(legend.position = c(0.11,0.91)) +
    #theme(plot.margin = unit(c(0, 0, 0, 0), "pt")) +
    theme(plot.title = element_blank()) +
    theme(legend.title = element_blank()) +
    theme(legend.background = element_rect(fill = NA))
  
  p_dis_1 <- ggplot(pcs_to_plot, aes(x = Axis.1, fill = Cohort)) +
    geom_density(alpha = 0.4, adjust = 1.5) +
    scale_fill_manual(
        values = c('COE' = 'darkred','DAP' = 'deepskyblue4')
        ) +
    scale_y_continuous(expand = c(0,0)) +
    theme_minimal() +
    xlab(NULL) +
    ylab(NULL) +
    theme(legend.position = 'none') +
    theme(axis.text = element_blank()) +
    theme(panel.grid = element_blank()) +
    theme(axis.line.x = element_line()) 
    #theme(plot.margin = unit(c(0, 0, 0, 0), "pt"))
  
  p_dis_2 <- ggplot(pcs_to_plot, aes(x = Axis.2, fill = Cohort)) +
    geom_density(alpha = 0.4, adjust = 1.5) +
    scale_fill_manual(
        values = c('COE' = 'darkred','DAP' = 'deepskyblue4')
        ) +
    scale_y_continuous(expand = c(0,0)) +
    theme_minimal() +
    xlab(NULL) +
    ylab(NULL) +
    theme(legend.position = 'none') +
    theme(axis.text = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(axis.title = element_blank()) +
    theme(axis.ticks.length = unit(0, "pt")) +
    theme(panel.grid = element_blank()) +
    theme(axis.line.y = element_line()) +
    coord_flip() +
    theme(plot.margin = unit(c(5, 0, 35, 0), "pt")) 
    
  p_left <- plot_grid(p_dis_1, p_pcoa, nrow = 2, rel_heights = c(1,4), align = 'v')
  p_right <- plot_grid(NULL, p_dis_2, nrow = 2, rel_heights = c(1,4))
  plot_grid(p_left, p_right, ncol = 2, rel_widths = c(4,1))
}
```

### All microbiome species, all dogs

```{r fig.height=4, fig.width=4.2}
pcs_all <- pcoa(bc_sp)
plot_pcoa_coe_vs_dap(pcs_all, metadata_all)
rm(pcs_all)
```

```{r warning=FALSE}
tmp <- get_permanova(
  disct_metadata = metadata_all %>% mutate(sample_id = paste(Cohort, sample_id, sep = '_')),
  dist_matrix = as.matrix(bc_sp) %>%
    as.data.frame() %>%
    rownames_to_column('sample_id'),
  dist_metric_name = 'Bray-Curtis',
  sample_id_column = 'sample_id',
  metadata_cols_to_analyze = c('Cohort'))
message('PERMANOVA p-value assessing overall differences in the microbiome: ', tmp$results$pvals[1])
rm(tmp)
```

### Common microbiome species only, all dogs

```{r fig.height=4, fig.width=4.5, eval = FALSE, echo = FALSE}
pcs_common_all <- pcoa(bc_sp_common)
p <- plot_pcoa_coe_vs_dap(pcs_common_all, metadata_all)
rm(pcs_common_all)
ggsave(file="SuppFigure1/SuppFigure1A.svg", plot=p, width=4.5, height=4)
p
```

# Barplots

```{r}
plot_taxonomy <- function(all_tax, rare_otu_cutoff = 0.01, 
                          plot_title = 'Taxonomy barplots', 
                          legend_title = 'Taxon', n_legend_cols = 2) {
  df <- all_tax %>%
    tidyr::pivot_longer(cols = -sample_id, names_to = 'taxon', values_to = 'relab') %>%
    # Re-normalize into relative abundance
    group_by(sample_id) %>%
    mutate(relab = relab / sum(relab)) %>%
    ungroup() %>%
    tidyr::separate(col = sample_id, 
                    into = c('Cohort','sample_id')) 
  
  # Group rare taxa together for plot simplicity
  rare_otus <- df %>%
    group_by(taxon) %>%
    summarise(mean_relab = mean(relab), .groups = 'drop') %>%
    filter(mean_relab < rare_otu_cutoff) %>%
    pull(taxon)
  
  df <- df %>%
    mutate(otu_grouped = ifelse(taxon %in% rare_otus, 'Others', taxon)) %>%
    group_by(sample_id, Cohort, otu_grouped) %>%
    summarise(relab = sum(relab), .groups = 'drop')
  
  # Slightly reorder to make the 'Others' category last
  taxa_list <- sort(unique(df$otu_grouped))
  taxa_list <- c(taxa_list[taxa_list != 'Others'], 'Others')
  df$otu_grouped <- factor(df$otu_grouped, levels = taxa_list)
  
  # Sort samples by the abundance of upmost taxon
  first_tax <- taxa_list[1]
  samp_order <- df %>%
    filter(otu_grouped == first_tax) %>%
    arrange(-relab) %>%
    pull(sample_id) 
  # Add samples with 0 abundance of this taxa
  samp_order <- c(samp_order, setdiff(unique(df$sample_id), samp_order))
  df$sample_id <- factor(df$sample_id, levels = samp_order)
  
  tmp_palette <- colorRampPalette(RColorBrewer::brewer.pal(9,name = 'RdYlGn'))(n_distinct(df$otu_grouped)-1)
  tmp_palette <- c(tmp_palette, 'grey') # For the 'Others' category
  
  message('Barplots include ',
          n_distinct(df$sample_id[df$Cohort == 'COE']), 
          ' COHELO samples and ',
          n_distinct(df$sample_id[df$Cohort == 'DAP']), 
          ' DAP samples')
  
  ggplot(df, aes(fill = otu_grouped, y = relab, x = sample_id)) + 
    geom_bar(position = "fill", stat = "identity") +
    scale_fill_manual(values = tmp_palette, name = legend_title) +
    theme_classic() +
    scale_y_continuous(expand = c(0,0)) +
    ggtitle(plot_title) +
    xlab('Sample') +
    ylab('Relative abundance') +
    guides(fill = guide_legend(ncol = n_legend_cols)) +
    facet_grid(. ~ Cohort, scales = "free_x", space = 'free_x') +
    theme(axis.text.x = element_blank()) +
    theme(plot.title = element_text(size = 13, hjust = 0.5)) +
    theme(legend.position = "bottom") +
    theme(legend.direction = "vertical") +
    theme(legend.text = element_text(size = 9)) +
    theme(legend.key.size = unit(9, "points")) +
    theme(panel.border = element_rect(colour = "black", fill = NA)) +
    theme(legend.key = element_rect(fill = NA, colour = "black")) +
    theme(axis.title = element_text(size = 13)) +
    theme(strip.background = element_blank()) +
    theme(strip.text = element_text(face = 'bold', size = 13))
}
```


## Phyla-level

Phyla with mean relative abundance (over all dogs in both datasets) < 2% are grouped into 'Others'.

```{r fig.width=10, fig.height=4, eval = FALSE}
plot_taxonomy(all_ph, rare_otu_cutoff = 0.01, plot_title = NULL)
```

## Genus-level

Genera with mean relative abundance (over all dogs in both datasets) < 0.5% are grouped into 'Others'.

```{r fig.width=10, fig.height=5}
p <- plot_taxonomy(
  all_ge, 
  rare_otu_cutoff = 0.01, 
  plot_title = NULL, 
  legend_title = 'Genus')
ggsave(file="Figure2/Figure2A.svg", plot=p, width=9, height=5)
p
```


## Averages - genus level

Each dot represents a genus. Rare genera removed.

```{r fig.height=2.5, fig.width=3.3}
tmp <- all_ge %>%
  tidyr::pivot_longer(
    cols = -sample_id, 
    names_to = 'taxon', 
    values_to = 'relab') %>%
  mutate(Cohort = gsub('_.*$', '', sample_id)) %>%
  # Get averages per cohort
  group_by(Cohort, taxon) %>%
  summarise(N = n(),
            relab_mean = mean(relab),
            relab_sd = sd(relab),
            .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = 'Cohort', 
              values_from = c('relab_mean', 'relab_sd', 'N'), 
              values_fill = 0) %>%
  # Completely remove rare taxa 
  filter(relab_mean_COE > 0.00001 & relab_mean_DAP > 0.00001) 

cor_result <- cor.test(tmp$relab_mean_COE, tmp$relab_mean_DAP, method = c("spearman"))
print(cor_result)

p <- ggplot(tmp, aes(x = relab_mean_COE, y = relab_mean_DAP)) +
  geom_point(size = 2, alpha = 0.4, color = 'black') +
  geom_smooth(method='lm', formula= y~x, color = 'chocolate3', linewidth = 1.1, alpha = 0.6, fill = 'blanchedalmond') +
  xlab('Mean genus abundance in COE') +
  ylab('Mean genus\nabundance in DAP') +
  scale_x_continuous(
    trans = log10_trans(), 
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) +
  scale_y_continuous(
    trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) +
  theme_classic() 

ggsave(file="Figure2/Figure2B.svg", plot=p, width=2.9, height=2)
p

rm(tmp)
```

## Averages - species level

Each dot represents a species. Rare species removed.

```{r fig.height=2.5, fig.width=3.3}
tmp <- all_sp %>%
  tidyr::pivot_longer(
    cols = -sample_id, 
    names_to = 'taxon', 
    values_to = 'relab') %>%
  mutate(Cohort = gsub('_.*$', '', sample_id)) %>%
  # Get averages per cohort
  group_by(Cohort, taxon) %>%
  summarise(N = n(),
            relab_mean = mean(relab),
            relab_sd = sd(relab),
            .groups = 'drop') %>%
  tidyr::pivot_wider(names_from = 'Cohort', 
              values_from = c('relab_mean', 'relab_sd', 'N'), 
              values_fill = 0) %>%
  # Completely remove rare species 
  filter(relab_mean_COE > 0.00001 & relab_mean_DAP > 0.00001) 

cor_result <- cor.test(tmp$relab_mean_COE, tmp$relab_mean_DAP, method = c("spearman"))
print(cor_result)
cor_label1 = "Spearman r: 0.69" # TODO: build automatically, was 0.86 with pearson
cor_label2 = "'p-value: ' < 10^{-10}" # TODO: build automatically

p <- ggplot(tmp, aes(x = relab_mean_COE, y = relab_mean_DAP)) +
  geom_point(size = 2, alpha = 0.4, color = 'black') +
  geom_smooth(method='lm', formula= y~x, color = 'chocolate3', linewidth = 1.1, alpha = 0.6, fill = 'blanchedalmond') +
  xlab('Mean species abundance in COE') +
  ylab('Mean species\nabundance in DAP') +
  annotate('text', x = 0.0001, y = 0.3, label=cor_label1, 
            hjust = 0, color = 'chocolate3', size = 3.7) +
  annotate('text', x = 0.0001, y = 0.1, label=cor_label2, 
            parse=TRUE, hjust = 0, color = 'chocolate3', size = 3.7) +
  scale_x_continuous(
    trans = log10_trans(),
    #limits = c(plot_lim_min, plot_lim_max),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) +
  scale_y_continuous(
    trans = log10_trans(),
    #limits = c(plot_lim_min, plot_lim_max),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) +
  theme_classic()
ggsave(file="SuppFigure1/SuppFigure1C.svg", plot=p, width=3.8, height=2.3)
p

rm(tmp, cor_label1, cor_label2, cor_result)
```

