---
title: "DAP dogs' microbiome shifts with age"
author: "Borenstein Lab"
date: "April 2025"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
---

```{r echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

Here we present the analysis that was performed to identify the association between the microbiome and age in dogs. The analysis was performed on the data from the Terra release 2023_12022024, and includes samples from 6 batches of dogs. The analysis includes both alpha and beta diversity, uniqueness, as well as differential abundance analysis and machine learning models to predict age based on microbiome features.
The analysis is Figure 3 in the paper (and also Figure 4C).

# Preparations

```{r message=FALSE, warning=FALSE}
# Preparations
library(readr)
library(ggplot2)
library(ggridges)
library(gridExtra)
library(tibble)
library(ggpubr)
library(tidyr)
library(dplyr)
library(themis)
library(ape)
library(cowplot)
library(vegan)
library(kableExtra)
library(stringr)
library(recipes)
library(ROSE)
library(combinat)
library(lightgbm)
library(caret)
library(ranger)
library(randomForest)
library(tidymodels)
options(scipen = 999)
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

## Read Terra data

```{r}

# Read HLES data
dog_and_owner <- read_csv(file.path(metadata_dir, "terra_release_2023_12022024", "DAP_2023_HLES_dog_owner_v1.0.csv"),
                         show_col_types = FALSE,
                         guess_max = 50000) %>%
  rename(dap_dog_id = dog_id) %>%
  mutate(dap_dog_id = as.character(dap_dog_id))

# Merge the sample_,ap and the HLES data
sample_map <- sample_map %>%
  inner_join(dog_and_owner, by="dap_dog_id")

# Add total_reads_before_filtering from qc_stats
sample_map <- sample_map %>% inner_join(qc_stats %>% select(sample_id, total_reads_before_filtering, duplication_rate), by="sample_id")

sample_map <- sample_map %>%
  filter(Breed_Size_Class_at_HLES != "Insufficient information to classify")

# Define the metadata columns to be used in the analysis as fixed effects and random effect
FIXED <- c("breed_size_numeric", "Sex_class", "total_reads_before_filtering", "duplication_rate", "hs_general_health", "weight_from_vet_or_survey_lbs") 
RANDOM <- c("pool") 

```

# Section 1: general aging-microbiome trends

## Age vs. alpha and beta-diversity

### Alpha diversity

Mixed effect model: age \~ alpha diversity + constant fixed effects

```{r fig.height=5,  fig.width=4, message=FALSE, warning=FALSE}

print("Get fixed effect model with shannon diversity")
alpha_div <- shannon %>% select(sample_id, shannon_sp)
res <- get_mixed_effect(alpha_div, 
                             sample_map, 
                             sample_id_column = 'sample_id', 
                             metadata_cols_to_analyze = c("age_at_stool_collection") , 
                             fdr_threshold = 0.05,
                             fixed_cols = FIXED,
                             random_cols = RANDOM)

res$plots # Figure  3A

```

### Beta diversity

```{r fig.height=3, fig.width=5, message=FALSE, warning=FALSE}


res <- get_distance_from_reference_group(
                      tmp_sample_map, 
                      bc, 
                      "Bray_Curtis",
                      metadata_col_to_analyze = "new_life_stage_at_stool_collection", 
                      reference_group_in_col = "Reference group young Adult",
                      sample_id_column = 'sample_id', 
                      metadata_cols_fix_effects = FIXED,
                      metadata_cols_random_effects = RANDOM,
                      fdr_threshold = 0.05,
                      colors = list("Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"),
                      ord = c("Young Adult", "Mature Adult", "Senior"))
print(res$plot + 
        labs(x = "Life state at stool collection") +
        labs(y = "Mean distance from young adults") +
        labs(title = paste(names(distances_dfs)[[i]])))
print(res$regression_results)
print(res$wilcoxon_resuls)



p <- res$plot + # Figure 3B
      labs(x = "Life state at stool collection") +
      labs(y = "Mean distance from young adults") 



```


```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}

pcoa_values_list <- list()
res_list <- list()


res <- get_permanova(disct_metadata = sample_map %>% filter(!is.na(weight_from_vet_or_survey_lbs)),
              dist_matrix = bc,
              dist_metric_name = "Bray_Curtis",
              sample_id_column = 'sample_id',
              metadata_cols_to_analyze = c("life_stage_at_stool_collection"),
              metadata_cols_fix_effects = c(RANDOM, FIXED),
              fdr_threshold = 0.1,
              ord=c("Puppy", "Young Adult", "Mature Adult", "Senior"),
              pcoa_fill = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"),
              add_centroid=TRUE,
              level=0.9
              )
p_threshold <- 0.05
results <- res$results
significant_res <- results[results$pvals < p_threshold, ]
if (dim(significant_res)[1] > 0){
  print(paste0('The following tests resulted in p-value < ', p_threshold, ":"))
  print(significant_res)
  p <- plot_grid(plotlist = res$plots, ncol = 2, nrow = 2, align = 'hv', axis = 'tblr')
  title <- ggdraw() + draw_label("PCoA of Permanova significant results", fontface='bold')
  print(plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)))
} else {
  print("No significat results to plot.")
}

```

Plot PCoA with age distribution:

```{r fig.height=5, fig.width=5.5, message=FALSE, warning=FALSE}

tmp_pcoa_values <- res$pcoa_values

# Get the pcoa plot
p_pcoa <- res_list$Bray_Curtis$plots$`life_stage_at_stool_collection_pc1&pc2` +
  labs(subtitle = NULL, title = NULL)

leg <- as_ggplot(get_legend(p_pcoa))

# Plot the pcoa with age distribution
p_pcoa <- p_pcoa + 
  theme(legend.title = element_blank()) +
  theme(legend.position = c(0.885,0.885)) +
  theme(legend.background = element_rect(fill = "transparent")) +
  scale_fill_manual(values = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"))

# Determine the x and y axis limits
xmin <- layer_scales(p_pcoa)$x$range$range[[1]]
xmax <- layer_scales(p_pcoa)$x$range$range[[2]]
ymin <- layer_scales(p_pcoa)$y$range$range[[1]]
ymax <- layer_scales(p_pcoa)$y$range$range[[2]]

# Set the number of bins and calculate the mean age for each bin
number_of_bins1 <- 50

# Set the number of bins and calculate the mean age for each bin
jumps1 <- (xmax - xmin) / number_of_bins1
df_mean_1 <- data.frame(bin = character(), mean_age = numeric())
for (i in 1:number_of_bins1){
  mean_age <- tmp_pcoa_values %>%
    select(Axis.1, age_at_stool_collection)  %>%
    filter(Axis.1 >= xmin + jumps1*(i-1) & Axis.1 < xmin + jumps1*i) %>%
    summarise(mean_age = mean(age_at_stool_collection, na.rm = TRUE))
  df_mean_1 <- rbind(df_mean_1, data.frame(bin=xmin+jumps1*(i-0.5),  mean_age = mean_age))
}

# Plot the mean age distribution of the 2nd axis
p_dis_1 <- ggplot(df_mean_1, aes(x = bin, y = mean_age)) +
  geom_bar(stat = "identity", fill = "#e0e1dd",  color = "grey70", alpha  = 0.2) +
   geom_smooth(data=df_mean_1,
              aes(x = bin, y = mean_age),
              method = "lm", se= FALSE, color = "firebrick4", size = 1) +
  xlab(NULL)+
  ylab("Mean age") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  xlim(xmin, xmax) 

#  Plot the mean age distribution of the 1st axis
number_of_bins2 <- 50
jumps2 <- (ymax - ymin) / number_of_bins2
df_mean_2 <- data.frame(bin = character(), mean_age = numeric())
for (i in 1:number_of_bins2){
  mean_age <- tmp_pcoa_values %>%
    select(Axis.2, age_at_stool_collection)  %>%
    filter(Axis.2 >= ymin + jumps2*(i-1) & Axis.2 < ymin + jumps2*i) %>%
    summarise(mean_age = mean(age_at_stool_collection, na.rm = TRUE))
  df_mean_2 <- rbind(df_mean_2, data.frame(bin=ymin+jumps2*(i-0.5),  mean_age = mean_age))
}
p_dis_2 <- ggplot(df_mean_2, aes(x = bin, y = mean_age)) +
  geom_bar(stat = "identity", fill = "#e0e1dd", color = "grey70", alpha  = 0.2) +
  geom_smooth(data=df_mean_2,
              aes(x = bin, y = mean_age),
              method = "lm", se= FALSE, color = "firebrick4", size = 1) +
  xlab(NULL) +
  ylab("Mean age") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  # theme(plot.margin = margin(0, 0, 0, 0)) +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  coord_flip() +
  xlim(ymin, ymax) 


# Create the final plot
p_left <- plot_grid(p_dis_1, p_pcoa, nrow = 2, rel_heights = c(0.7,4), align = 'v')
p_right <- plot_grid(NULL, p_dis_2, nrow = 2, rel_heights = c(0.7,4))
p <- plot_grid(p_left, p_right, ncol = 2, rel_widths = c(4,0.7))

p # Figure 3C
```


#### Uniqueness

```{r fig.height=2.5, fig.width=2, message=FALSE, warning=FALSE}


min_non_zero <- function(x) {
  min_val <- min(x[x > 0], na.rm = TRUE)
  ifelse(is.infinite(min_val), NA, min_val)
}


# Apply the function across rows to get the minimal non-zero value for each row
uni <- bc %>%
  rowwise() %>%
  mutate(uniqueness = min_non_zero(c_across(-sample_id)))

# Select only the sample_id and uniqueness columns to create the new dataframe
uni <- select(uni, sample_id, uniqueness)

res <- get_mixed_effect(uni, 
                        sample_map %>% filter(!is.na(weight_from_vet_or_survey_lbs)), 
                        sample_id_column = 'sample_id', 
                        metadata_cols_to_analyze = c("age_at_stool_collection") , 
                        fdr_threshold = 0.05,
                        fixed_cols = FIXED,
                        random_cols = RANDOM)
if (length(res$plots) > 0){
  print(res$plots[[1]] + labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value"))  )
} else {
  print("No significant correlations to plot.")
}
  
res$plots[[1]] + # Figure 3D
  labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value")) + 
  labs(title = names(distances_dfs)[[i]])
  

```

# Section 2: specific species shifts - aging markers

```{r fig.height=4, fig.width=5.5, message=FALSE, warning=FALSE}

res <- run_continuous_maaslin(feat_table = ra_kraken_sp,
                                metadata =  sample_map %>% 
                                 filter(!is.na(weight_from_vet_or_survey_lbs)) %>%  
                                 mutate(breed_size_numeric = as.numeric(breed_size_numeric)), 
                                sample_id_column = 'sample_id', 
                                metadata_cols_to_analyze = c("age_at_stool_collection"), 
                                metadata_cols_fix_effects = c(FIXED),
                                metadata_cols_rand_effects = RANDOM,
                                fdr_threshold = 0.05,
                                output_dir = "da",
                                transformation_method = "LOG",
                                keep_maaslin_plots = FALSE,
                                category_name_map=NULL,
                                logscale_y = TRUE,
                                return_maaslin_plots = FALSE)

# Load the GTDB taxonomy to get the genus and phylum per each species
gtdb <- samples_metadata <- read_tsv(
  file.path(metadata_dir, "gtdb", "bac120_taxonomy_r207.tsv"),show_col_types = FALSE, col_names=FALSE) %>%
  rename(taxonomy = X2) %>%   
  separate(taxonomy, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";") %>%
  mutate(
    Domain = sub("d__", "", Domain),
    Phylum = sub("p__", "", Phylum),
    Class = sub("c__", "", Class),
    Order = sub("o__", "", Order),
    Family = sub("f__", "", Family),
    Genus = sub("g__", "", Genus),
    Species = sub("s__", "", Species)
  )

# Plot species with significant association with age
p <- res$all_results %>%
  filter(value == "age_at_stool_collection") %>%
  mutate(Genus = str_extract(feature, "^[^.]+")) %>%
  left_join(gtdb %>% select(Genus, Phylum) %>% distinct(), by = "Genus") %>%
    mutate(Genus = case_when(
    Genus %in% c("Prevotella", "Holdemanella") ~ Genus,
    TRUE ~ "other"
  )) %>%
  ggplot(aes(x = coef, y = -log(FDR), fill=Genus, color=Genus)) +
  geom_point( alpha = 0.7, shape = 21, size = 2.5) +
  geom_hline(yintercept = -log(0.05), linetype = "dashed", color = "red") +  
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + 
  labs(x = "Regression coefficient", y = "-log(FDR)") +  # Axis labels
  # Color specific genus
  scale_fill_manual(values = c("Prevotella" = "#f7b267",  "Holdemanella" = "#c9184a", "other" = "grey80"), 
                    breaks = c("Prevotella", "Holdemanella", "C")) +
  scale_color_manual(values = c("Prevotella" = "grey20",  "Holdemanella" = "grey20", "other" = "grey20"),
                     breaks = c("Prevotella", "Holdemanella", "C")) +
  theme_classic()

p # Figure 3E
```

# Section 3: predict age based on microbiome features

```{r}
get_predictability_regression_for_age_pred <- function(feat_table, 
                                                       metadata, 
                                                       sample_id_column, 
                                                       vars_to_analyze, 
                                                       ml_method = 'rf',
                                                       fdr_threshold = 0.1,
                                                       corr_threshold = 0.3,
                                                       corr_method = "pearson",
                                                       n_ml_repeats = 10,
                                                       n_ml_folds = 5,
                                                       feat_importance = 'none',
                                                       rand_seed = 1111,
                                                       quiet = FALSE) {
  # Required libraries loading
  require(rsample)
  require(dplyr)
  require(glmnet)
  require(ranger)
  require(metap) 
  
  # Argument validations 
  if(! all(vars_to_analyze %in% names(metadata)))
    stop("Invalid 'vars_to_analyze' argument. Some of the variables provided are missing from the metadata table")
  if(is.null(vars_to_analyze))
    stop("Invalid 'vars_to_analyze' argument. Cannot be NULL")
  if(! ml_method %in% c('rf'))
    stop("Invalid 'ml_method' argument. Only 'rf' supported.")
  if (! corr_method %in% c('pearson', 'spearman'))
    stop("Invalid 'corr_method' argument. Only 'pearson' and 'spearman' supported.")
  
  # Main calculations
  n_samples <- nrow(feat_table)
  
  # Generate folds for cross-validation
  set.seed(rand_seed)
  folds <- vfold_cv(
    data.frame(sample_num = 1:n_samples),
    v = n_ml_folds,
    repeats = n_ml_repeats
  )
  
  # Collect evaluation stats here
  cv_correlations <- data.frame()
  cv_feat_imp <- data.frame()
  
  # Fix names of features (causes errors in some models)
  names(feat_table) <- make.names(names(feat_table))
  
  # Iterate over variables
  for (measure in vars_to_analyze) {
    if (!quiet) message("Training model for variable: ", measure)
    
    # Prepare temporary table for training (features + label only)
    tmp <- feat_table %>% 
      left_join(
        metadata[,c(sample_id_column, measure)] %>% 
          rename(Label = 2), 
        by = sample_id_column)

    pred_per_fold_list <- list()
    
    # Cross validation loop
    for (i in 1:nrow(folds)) {
      fold_id <- paste(folds$splits[[i]]$id, collapse = " ")
      
      # Get train/test samples of this fold
      train_samples <- folds$splits[[i]]$in_id
      train_data <- tmp[train_samples,] %>% filter(!is.na(Label))
      test_data <- tmp[-train_samples,] %>% filter(!is.na(Label))
      
      test_data_with_sample_id <- test_data
      train_data <- train_data %>%
        select(-any_of(c(sample_id_column)))
      test_data <- test_data %>%
        select(-any_of(c(sample_id_column)))
      
      # Train RF
      rf_model <- ranger(Label ~ ., data = train_data, importance = ifelse(feat_importance=='altmann','permutation','none'), seed = rand_seed)
      if (feat_importance != 'none') {
        feat_imp <- importance_pvalues(rf_model, method="altmann", formula = Label ~ ., data = train_data, seed = rand_seed)
      }
      
      # Make prediction on held out samples
      preds <- predict(rf_model, test_data %>% select(-Label))$predictions
      
      
      # Calculate model performance for this fold
      tmp_eval <- data.frame(preds = preds, true_label = test_data$Label, sample_id = test_data_with_sample_id$sample_id)  
      
      pred_per_fold_list[[i]] <- tmp_eval
      
      cv_correlations <- bind_rows(
        cv_correlations,
        data.frame(
          measure = measure,
          fold_id = fold_id,
          test_cor = cor(tmp_eval$preds, tmp_eval$true_label, method = corr_method),
          test_cor_p = cor.test(tmp_eval$preds, 
                                tmp_eval$true_label, 
                                alternative = 'greater', # Check positive association
                                method = corr_method,
                                exact = FALSE)$p.value
        )
      )
      
      if (feat_importance != 'none')
        cv_feat_imp <- bind_rows(
          cv_feat_imp,
          as.data.frame(feat_imp) %>% 
            tibble::rownames_to_column('feature') %>%
            mutate(measure = measure, fold_id = fold_id)
        )
    }
  }
  
  sumlog2 <- function(ps) {ps <- ps[!is.na(ps)]; return(sumlog(ps)$p)}
  
  # Summarise over folds
  summary_correlations <- cv_correlations %>%
    group_by(measure) %>%
    summarise(n_folds = n(),
              mean_cor = mean(test_cor, na.rm = TRUE),
              sd_cor = sd(test_cor, na.rm = TRUE),
              p = sumlog2(test_cor_p),
              .groups = "drop") %>%
    mutate(fdr = p.adjust(p)) %>%
    mutate(Significant = fdr < fdr_threshold & mean_cor > corr_threshold)
  
  # Summarize feature importance (if requested by user)
  summary_feat_importance <- data.frame()
  if (feat_importance != 'none')
    summary_feat_importance <- cv_feat_imp %>%
      group_by(measure, feature) %>%
      summarise(n_folds = n(),
                mean_feat_importance = mean(importance, na.rm = TRUE),
                sd_feat_importance = sd(importance, na.rm = TRUE),
                p = sumlog2(pvalue),
                .groups = "drop") %>%
      mutate(fdr = p.adjust(p))
  
  if (! any(summary_correlations$Significant)) {
    message('No significantly well-predicted metadata measures')
  } else {
    message('The following measures were well-predicted by the microbiome using a ', toupper(ml_method), ' model:')
    message(paste(summary_correlations$measure[summary_correlations$Significant], collapse = ', '))
  }
  
  return(list(results = summary_correlations, feat_importance = summary_feat_importance, cv_cor=cv_correlations, fold_preds=pred_per_fold_list))
}

```

Train the model with microbiome features + confounders and only confounders:

```{r}

PRED_FIXED <- c("breed_size_numeric", "Sex_bool", "pool", "total_reads_before_filtering",
                "duplication_rate", "weight_from_vet_or_survey_lbs")

# microbiome features + confounders

tmp_sample_map <- sample_map %>% 
  filter(!is.na(weight_from_vet_or_survey_lbs))  %>%
  filter(!is.na(breed_size_numeric))  %>%
  filter(hs_general_health %in% c(1,2)) %>%
  mutate(Sex_bool = case_when(
    grepl("Female", Sex_class) ~ "female",
    grepl("Male", Sex_class) ~ "male",
    TRUE ~ Sex_class # This is a catch-all in case there are unexpected values
  ))

tmp_ra_kraken_sp <- ra_kraken_sp %>% 
  inner_join(tmp_sample_map %>% select(PRED_FIXED, sample_id), by="sample_id") 

rf_res_mic <- get_predictability_regression_for_age_pred(tmp_ra_kraken_sp, 
                                                         tmp_sample_map, 
                                                         "sample_id", 
                                                         "age_at_stool_collection", 
                                                         ml_method = 'rf',
                                                         fdr_threshold = 1,
                                                         corr_threshold = 0,
                                                         corr_method = "pearson",
                                                         n_ml_repeats = 1, 
                                                         n_ml_folds = 10, 
                                                         feat_importance = 'none',
                                                         rand_seed = 84,
                                                         quiet = FALSE)

# only confounders

confounders <- tmp_sample_map %>% select(sample_id, PRED_FIXED) %>%
  mutate(breed_size_numeric = as.numeric(breed_size_numeric))

rf_res_con <- get_predictability_regression3(confounders, 
                                          tmp_sample_map, 
                                          "sample_id", 
                                          "age_at_stool_collection", 
                                          ml_method = 'rf',
                                          fdr_threshold = 1,
                                          corr_threshold = 0,
                                          corr_method = "pearson",
                                          n_ml_repeats = 1, 
                                          n_ml_folds = 10, 
                                          feat_importance = 'none',
                                          rand_seed = 84,
                                          quiet = FALSE)


final_df_mic <- do.call(rbind, lapply(seq_along(rf_res_mic$fold_preds), function(i) {
  df <- rf_res_mic$fold_preds[[i]]
  df$index <- i
  df$data <- "microbiome + confounders"
  return(df)
}))

final_df_con <- do.call(rbind, lapply(seq_along(rf_res_con$fold_preds), function(i) {
  df <- rf_res_con$fold_preds[[i]]
  df$index <- i
  df$data <- "confounders"
  return(df)
}))

merged_df <- rbind(final_df_mic, final_df_con)

```

Plot the results of the models:

```{r fig.height=4, fig.width=4, message=FALSE, warning=FALSE}

library(ggpmisc)

# Calculate correlations per facet
correlations <- merged_df %>%
  filter(data == "microbiome + confounders") %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds)) %>%
  mutate(data="microbiome + confounders")
correlations2 <- merged_df %>%
  filter(data == "confounders") %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds)) %>%
  mutate(data="confounders")

merged_correlations <- rbind(correlations, correlations2)

p <- ggplot(merged_correlations, aes(x = data, y = correlation)) +
  geom_boxplot(aes(fill=data), alpha=0.5, show.legend = FALSE) +
  geom_point(size=2.5, alpha=0.5) + #, aes(color=as.factor(index))
  labs(x = "Training data", y = "Perason correlation") + # , color = "Fold"
  # scale_color_manual(values = c("#012a4a", "#013a63", "#01497c", "#014f86", "#2a6f97", "#2c7da0", "#468faf", "#61a5c2", "#89c2d9", "#a9d6e5")) +
  scale_fill_manual(values = c("#b5e48c", "#89c2d9")) +
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "lightgray", size = 0.25),  # Major grid lines
    panel.grid.minor = element_line(color = "lightgray", size = 0.25),  # Minor grid lines
    panel.grid.major.x = element_blank(),  # Remove vertical grid lines
    panel.grid.minor.x = element_blank(),  # Remove minor vertical grid lines
  ) +
  scale_x_discrete(labels = c("microbiome + confounders" = "Microbiome + confounders", "confounders" = "Confounders"))  # Manually set x-axis labels

print(p) # Figure 3G

```

Plot the correlation between the real and predicted age per fold (for the model that used microbiome features + confounders):

```{r fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Load necessary packages
library(ggplot2)
library(dplyr)
library(ggpubr)

data <- merged_df %>% filter(data == "microbiome + confounders")

# Calculate Spearman correlations for each index
correlations <- data %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds, method = "pearson")) %>%
  mutate(label = paste(index))#, ": Pearson r =", round(correlation, 3)))

# Convert correlations to a named vector for legend labels
legend_labels <- setNames(correlations$label, as.character(correlations$index))

# Create the scatter plot with regression lines
p <- ggplot(data, aes(x = true_label, y = preds)) +
  geom_point(size=1.25, alpha=0.3) +
  geom_smooth(aes(color = as.factor(index)), method = "lm", se = FALSE, size=0.5, alpha=0.3) +
  labs(color = "Fold") +
  xlab("Real age") +
  ylab("Predicted age") +
  scale_color_manual(values = c("#012a4a", "#013a63", "#01497c", "#014f86", "#2a6f97", "#2c7da0", "#468faf", "#61a5c2", "#89c2d9", "#a9d6e5"), 
                     labels = legend_labels) +
  theme_classic()

# Display the plot
print(p) # Figure 3F

```

Import clinical lab tests

```{r}

# Import CBC lab test
cbc_lab_test <- read_csv(file.path(metadata_dir,"terra_release_2023_12022024","DAP_2023_SamplesResults_CBC_v1.0.csv"), 
                         show_col_types = FALSE, na = c('NA'))%>%
  # filter to precision 1 dogs with metagenomics
  filter(Sample_Year == "precision_1")%>%
  rename(dap_dog_id = dog_id )%>%
  mutate(dap_dog_id = as.character(dap_dog_id))%>%
  # filter non-numeric and NA columns 
  select(where(~ !all(is.na(.))))%>%
  select(-c(krt_cbc_abs_bands, krt_cbc_pcv, krt_cbc_plt_quant, krt_cbc_plt_morph, krt_cbc_plt_morph_num, krt_cbc_rbc_morph, krt_cbc_rbc_morph_2, krt_cbc_rbc_morph_3, krt_cbc_rbc_morph_num, krt_cbc_rbc_morph_num_2, krt_cbc_wbc_morph, krt_cbc_wbc_morph_num, krt_cbc_hemoparasites, krt_cbc_test_comments, krt_cbc_abs_metamyelocytes, krt_cbc_abs_other_cells, krt_cbc_rel_metamyelocytes, krt_cbc_rel_bands, krt_cbc_rel_basophils, "krt_cbc_pct","krt_cbc_plt_morph_2", "krt_cbc_plt_morph_num_2", "krt_cbc_nucleated_rbcs" , "krt_cbc_rbc_morph_num_3"))%>%   select(-matches("krt_cbc_rel*"))%>%
  # add metagenomic sample id
  left_join(select(sample_map, sample_id, dap_dog_id))%>%
  
cbc_numeric_columns <- cbc_lab_test %>%
  select_if(is.numeric) %>%
  names()

# Import CP lab test
cp_lab_test <- read_csv(file.path(metadata_dir,"terra_release_2023_12022024","DAP_2023_SamplesResults_ChemistryPanel_v1.0.csv"), show_col_types = FALSE, na = c('NA'))%>%
  # filter to precision 1 dogs with metagenomics
  filter(Sample_Year == "precision_1")%>%
  rename(dap_dog_id = dog_id )%>%
  mutate(dap_dog_id = as.character(dap_dog_id))%>%
  select(where(~ !all(is.na(.))))%>%
  select(-krt_cp_test_comments)%>%
  # add metagenomic sample id
  left_join(select(sample_map, sample_id, dap_dog_id))%>%
  
cp_numeric_columns <- cp_lab_test %>%
  select_if(is.numeric) %>%
  names()

#Import urine lab test - mostly char
urine_lab_test <- read_csv(file.path(metadata_dir,"terra_release_2023_12022024","DAP_2023_SamplesResults_Urine_v1.0.csv"), show_col_types = FALSE, na = c('NA'))%>%
  filter(Sample_Year == "precision_1")%>%
  rename(dap_dog_id = dog_id )%>%
  # filter(dap_dog_id %in% dap_id_subset$dap_dog_id)%>%
  mutate(dap_dog_id = as.character(dap_dog_id))%>%
  left_join(select(sample_map, sample_id, dap_dog_id))

```


```{r fig.height=7, fig.width=4, message=FALSE, warning=FALSE}

library(effectsize)

lab_test_results <- cbc_lab_test %>% select(sample_id, all_of(cbc_numeric_columns)) %>% drop_na(sample_id) %>% 
  inner_join(cp_lab_test %>% select(sample_id, all_of(cp_numeric_columns)) %>% drop_na(sample_id), by="sample_id") %>% 
  inner_join(urine_lab_test %>% 
               select(sample_id, krt_urine_ph, krt_urine_sg) %>% # 
               drop_na(sample_id) %>%
               mutate(krt_urine_sg = ifelse(krt_urine_sg == ">1.060", "1.060", krt_urine_sg)) %>%
               mutate(krt_urine_sg = as.numeric(krt_urine_sg)), by="sample_id") # %>%

# Impute missing values with half of the minimum value in the column
replace_na_with_half_min <- function(column) {
  min_val <- min(column, na.rm = TRUE)
  column[is.na(column)] <- min_val / 2
  return(column)
}

# Apply the function to each column in the data frame
lab_test_results <- lab_test_results %>%   mutate(across(where(is.numeric), replace_na_with_half_min))

model <- lm(preds ~ true_label, data = final_df_mic)
residuals <- resid(lm(preds ~ true_label, data = final_df_mic))

residuals_df <- final_df_mic %>%
  mutate(residuals = resid(lm(preds ~ true_label, data = final_df_mic))) %>%
  inner_join(lab_test_results, by="sample_id") 
  
regressors <- lab_test_results %>% select (-sample_id) %>% colnames()

final_pctexp <- data.frame("regressor" = c(sort(regressors)))
formula_string <- paste("residuals", "~", paste(regressors, collapse = " + "))
model <- lm(formula = as.formula(formula_string), data = residuals_df)
an <- car::Anova(model, type = 2)
andf <- data.frame(regressor = rownames(an), pval = an$"Pr(>F)", ss=an$"Sum Sq") %>% filter(regressor != "Residuals")
e <- eta_squared(an, partial = TRUE) # partial = TRUE is the default
residuals_df <- data.frame(regressor = e$Parameter, pctexp = e$Eta2_partial) 

residuals_df <- residuals_df %>% 
  mutate(regressor = factor(residuals_df$regressor, levels=regressors)) %>%
  separate(regressor, into = c("krt", "lab_test_category"), sep = "_", remove = FALSE) %>%
  select(-krt) %>%
  inner_join(andf, by="regressor") %>%
  mutate(pctexp = if_else(is.na(pctexp), 0, pctexp)) %>%
  mutate(fdr = p.adjust(pval, method = "fdr")) %>%
  mutate(significance = ifelse(fdr < 0.1, "*", "")) %>%
  left_join(codebook, by = c("regressor" = "Variable")) %>%
  mutate(SurveyText = str_replace(SurveyText, " Value", "")) %>%
  mutate(lab_test_category = case_when(
    lab_test_category == "cbc" ~ "Complete blood count",
    lab_test_category == "cp" ~ "Chemistry profile",
    lab_test_category == "urine" ~ "Urine",
    TRUE ~ lab_test_category  # Keeps any other values unchanged
  ))


```

```{r fig.height=7, fig.width=9, message=FALSE, warning=FALSE}

library(effectsize)

lab_test_results <- cbc_lab_test %>% select(sample_id, all_of(cbc_numeric_columns)) %>% drop_na(sample_id) %>% 
  inner_join(cp_lab_test %>% select(sample_id, all_of(cp_numeric_columns)) %>% drop_na(sample_id), by="sample_id") %>% 
  inner_join(urine_lab_test %>% 
               select(sample_id, krt_urine_ph, krt_urine_sg) %>% # 
               drop_na(sample_id) %>%
               mutate(krt_urine_sg = ifelse(krt_urine_sg == ">1.060", "1.060", krt_urine_sg)) %>%
               mutate(krt_urine_sg = as.numeric(krt_urine_sg)), by="sample_id") # %>%
  # select(-krt_cbc_total_wbcs)


# Impute missing values with half of the minimum value in the column
replace_na_with_half_min <- function(column) {
  min_val <- min(column, na.rm = TRUE)
  column[is.na(column)] <- min_val / 2
  return(column)
}

# Apply the function to each column in the data frame
lab_test_results <- lab_test_results %>%   mutate(across(where(is.numeric), replace_na_with_half_min))

age_df <- sample_map %>% 
  select(sample_id, age_at_stool_collection) %>% 
  inner_join(lab_test_results, by="sample_id") 

regressors <- lab_test_results %>% select (-sample_id) %>% colnames()

final_pctexp <- data.frame("regressor" = c(sort(regressors)))
formula_string <- paste("age_at_stool_collection", "~", paste(regressors, collapse = " + "))
model <- lm(formula = as.formula(formula_string), data = age_df)
an <- car::Anova(model, type = 2)
andf <- data.frame(regressor = rownames(an), pval = an$"Pr(>F)", ss=an$"Sum Sq") %>% filter(regressor != "Residuals")
e <- eta_squared(an, partial = TRUE) # partial = TRUE is the default
real_age_df <- data.frame(regressor = e$Parameter, pctexp = e$Eta2_partial) 


real_age_df <- real_age_df %>% 
  mutate(regressor = factor(real_age_df$regressor, levels=regressors)) %>%
  separate(regressor, into = c("krt", "lab_test_category"), sep = "_", remove = FALSE) %>%
  select(-krt) %>%
  inner_join(andf, by="regressor") %>%
  mutate(pctexp = if_else(is.na(pctexp), 0, pctexp)) %>%
  mutate(fdr = p.adjust(pval, method = "fdr")) %>%
  mutate(significance = ifelse(fdr < 0.05, "*", "")) %>%
  left_join(codebook, by = c("regressor" = "Variable")) %>%
  mutate(SurveyText = str_replace(SurveyText, " Value", "")) %>%
  mutate(lab_test_category = case_when(
    lab_test_category == "cbc" ~ "Complete blood count",
    lab_test_category == "cp" ~ "Chemistry profile",
    lab_test_category == "urine" ~ "Urine",
    TRUE ~ lab_test_category  # Keeps any other values unchanged
  ))

real_age_df <- residuals_df %>%
  mutate(param = "Metagenomic clock residuals") %>%
  rbind(real_age_df %>% mutate(param = "Age")) %>%
  mutate(lab_test_category = case_when(
      lab_test_category %in% c("cbc", "cp") ~ toupper(lab_test_category),
      lab_test_category == "urine" ~ "Urine",
      TRUE ~ lab_test_category  # Keep unchanged if none of the conditions match
    )) 

# Arrange the data by regressor order where param == "age"
real_age_df_ordered <- real_age_df %>%
  filter(param == "Age") %>%
  arrange(pctexp)  # Arrange by pctexp in descending order for example


p <- ggplot(data = real_age_df %>% mutate(pctexp = pctexp), aes(y = pctexp, x = factor(SurveyText, levels = real_age_df_ordered$SurveyText), fill = lab_test_category)) +
  geom_bar(stat = "identity", alpha=0.7, color="black") +
  theme_classic() +
  geom_text(data=real_age_df,aes(y = pctexp + 0.001, x = factor(SurveyText, levels = edf2_ordered$SurveyText), label = significance)) +
  theme(axis.text.y = element_text(angle = 0, hjust = 1), legend.position = "none") +  # Adjust angle as needed
  ylab("Percentage explained (age)") +
  xlab("Regressor") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip() +
  scale_fill_manual(values = list("Complete blood count" = "#f95738", "Chemistry profile" = "#ee964b", "Urine" = "#f4d35e")) +
  facet_grid(lab_test_category ~factor(param, levels=c('Age', 'Metagenomic clock residuals')), scales = "free", space = "free", switch = "y") +
  scale_y_continuous(trans = "reverse") # Reverse the y-axis


print(p) # Figure 4C

```
 