---
title: "DAP dogs' microbiome shifts with age"
author: "Borenstein Lab"
date: "April 2025"
output:
  html_document:
    css: custom_notebook_formatting.css
    toc: true
    toc_depth: 3
    df_print: paged
    code_folding: hide
---

```{r echo = FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center')
```

Here we present the analysis that was performed to identify the association between the microbiome and age in dogs. The analysis was performed on the data from the Terra release 2023_12022024, and includes samples from 6 batches of dogs. The analysis includes both alpha and beta diversity, uniqueness, as well as differential abundance analysis and machine learning models to predict age based on microbiome features.
The analysis is Figure 3 in the paper (and also Figure 4C).

# Preparations

```{r message=FALSE, warning=FALSE}
# Preparations
library(readr)
library(ggplot2)
library(ggridges)
library(gridExtra)
library(tibble)
library(ggpubr)
library(tidyr)
library(dplyr)
library(themis)
library(ape)
library(cowplot)
library(vegan)
library(kableExtra)
library(stringr)
library(recipes)
library(ROSE)
library(combinat)
library(lightgbm)
library(caret)
library(ranger)
library(randomForest)
library(tidymodels)
options(scipen = 999)
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

## Read Terra data

```{r}

# Read HLES data
dog_and_owner <- read_csv(file.path(metadata_dir, "terra_release_2023_12022024", "DAP_2023_HLES_dog_owner_v1.0.csv"),
                         show_col_types = FALSE,
                         guess_max = 50000) %>%
  rename(dap_dog_id = dog_id) %>%
  mutate(dap_dog_id = as.character(dap_dog_id))

# Merge the sample_,ap and the HLES data
sample_map <- sample_map %>%
  inner_join(dog_and_owner, by="dap_dog_id")

# Add total_reads_before_filtering from qc_stats
sample_map <- sample_map %>% inner_join(qc_stats %>% select(sample_id, total_reads_before_filtering, duplication_rate), by="sample_id")

sample_map <- sample_map %>%
  filter(Breed_Size_Class_at_HLES != "Insufficient information to classify")

# Define the metadata columns to be used in the analysis as fixed effects and random effect
FIXED <- c("breed_size_numeric", "Sex_class", "total_reads_before_filtering", "duplication_rate", "hs_general_health", "weight_from_vet_or_survey_lbs") 
RANDOM <- c("pool") 

```

# Section 1: general aging-microbiome trends

## Age vs. alpha and beta-diversity

### Alpha diversity

Mixed effect model: age \~ alpha diversity + constant fixed effects

```{r fig.height=5,  fig.width=4, message=FALSE, warning=FALSE}

print("Get fixed effect model with shannon diversity")
alpha_div <- shannon %>% select(sample_id, shannon_sp)
res <- get_mixed_effect(alpha_div, 
                             sample_map, 
                             sample_id_column = 'sample_id', 
                             metadata_cols_to_analyze = c("age_at_stool_collection") , 
                             fdr_threshold = 0.05,
                             fixed_cols = FIXED,
                             random_cols = RANDOM)

res$plots # Figure  3A

```

### Beta diversity

```{r fig.height=3, fig.width=5, message=FALSE, warning=FALSE}


res <- get_distance_from_reference_group(
                      tmp_sample_map, 
                      bc, 
                      "Bray_Curtis",
                      metadata_col_to_analyze = "new_life_stage_at_stool_collection", 
                      reference_group_in_col = "Reference group young Adult",
                      sample_id_column = 'sample_id', 
                      metadata_cols_fix_effects = FIXED,
                      metadata_cols_random_effects = RANDOM,
                      fdr_threshold = 0.05,
                      colors = list("Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"),
                      ord = c("Young Adult", "Mature Adult", "Senior"))
print(res$plot + 
        labs(x = "Life state at stool collection") +
        labs(y = "Mean distance from young adults") +
        labs(title = paste(names(distances_dfs)[[i]])))
print(res$regression_results)
print(res$wilcoxon_resuls)



p <- res$plot + # Figure 3B
      labs(x = "Life state at stool collection") +
      labs(y = "Mean distance from young adults") 



```


```{r fig.height=10, fig.width=12, message=FALSE, warning=FALSE}

pcoa_values_list <- list()
res_list <- list()


res <- get_permanova(disct_metadata = sample_map %>% filter(!is.na(weight_from_vet_or_survey_lbs)),
              dist_matrix = bc,
              dist_metric_name = "Bray_Curtis",
              sample_id_column = 'sample_id',
              metadata_cols_to_analyze = c("life_stage_at_stool_collection"),
              metadata_cols_fix_effects = c(RANDOM, FIXED),
              fdr_threshold = 0.1,
              ord=c("Puppy", "Young Adult", "Mature Adult", "Senior"),
              pcoa_fill = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"),
              add_centroid=TRUE,
              level=0.9
              )
p_threshold <- 0.05
results <- res$results
significant_res <- results[results$pvals < p_threshold, ]
if (dim(significant_res)[1] > 0){
  print(paste0('The following tests resulted in p-value < ', p_threshold, ":"))
  print(significant_res)
  p <- plot_grid(plotlist = res$plots, ncol = 2, nrow = 2, align = 'hv', axis = 'tblr')
  title <- ggdraw() + draw_label("PCoA of Permanova significant results", fontface='bold')
  print(plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)))
} else {
  print("No significat results to plot.")
}

```

Plot PCoA with age distribution:

```{r fig.height=5, fig.width=5.5, message=FALSE, warning=FALSE}

tmp_pcoa_values <- res$pcoa_values

# Get the pcoa plot
p_pcoa <- res_list$Bray_Curtis$plots$`life_stage_at_stool_collection_pc1&pc2` +
  labs(subtitle = NULL, title = NULL)

leg <- as_ggplot(get_legend(p_pcoa))

# Plot the pcoa with age distribution
p_pcoa <- p_pcoa + 
  theme(legend.title = element_blank()) +
  theme(legend.position = c(0.885,0.885)) +
  theme(legend.background = element_rect(fill = "transparent")) +
  scale_fill_manual(values = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  "Mature Adult" = "#92C5DE", "Senior" = "#0571B0"))

# Determine the x and y axis limits
xmin <- layer_scales(p_pcoa)$x$range$range[[1]]
xmax <- layer_scales(p_pcoa)$x$range$range[[2]]
ymin <- layer_scales(p_pcoa)$y$range$range[[1]]
ymax <- layer_scales(p_pcoa)$y$range$range[[2]]

# Set the number of bins and calculate the mean age for each bin
number_of_bins1 <- 50

# Set the number of bins and calculate the mean age for each bin
jumps1 <- (xmax - xmin) / number_of_bins1
df_mean_1 <- data.frame(bin = character(), mean_age = numeric())
for (i in 1:number_of_bins1){
  mean_age <- tmp_pcoa_values %>%
    select(Axis.1, age_at_stool_collection)  %>%
    filter(Axis.1 >= xmin + jumps1*(i-1) & Axis.1 < xmin + jumps1*i) %>%
    summarise(mean_age = mean(age_at_stool_collection, na.rm = TRUE))
  df_mean_1 <- rbind(df_mean_1, data.frame(bin=xmin+jumps1*(i-0.5),  mean_age = mean_age))
}

# Plot the mean age distribution of the 2nd axis
p_dis_1 <- ggplot(df_mean_1, aes(x = bin, y = mean_age)) +
  geom_bar(stat = "identity", fill = "#e0e1dd",  color = "grey70", alpha  = 0.2) +
   geom_smooth(data=df_mean_1,
              aes(x = bin, y = mean_age),
              method = "lm", se= FALSE, color = "firebrick4", size = 1) +
  xlab(NULL)+
  ylab("Mean age") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  xlim(xmin, xmax) 

#  Plot the mean age distribution of the 1st axis
number_of_bins2 <- 50
jumps2 <- (ymax - ymin) / number_of_bins2
df_mean_2 <- data.frame(bin = character(), mean_age = numeric())
for (i in 1:number_of_bins2){
  mean_age <- tmp_pcoa_values %>%
    select(Axis.2, age_at_stool_collection)  %>%
    filter(Axis.2 >= ymin + jumps2*(i-1) & Axis.2 < ymin + jumps2*i) %>%
    summarise(mean_age = mean(age_at_stool_collection, na.rm = TRUE))
  df_mean_2 <- rbind(df_mean_2, data.frame(bin=ymin+jumps2*(i-0.5),  mean_age = mean_age))
}
p_dis_2 <- ggplot(df_mean_2, aes(x = bin, y = mean_age)) +
  geom_bar(stat = "identity", fill = "#e0e1dd", color = "grey70", alpha  = 0.2) +
  geom_smooth(data=df_mean_2,
              aes(x = bin, y = mean_age),
              method = "lm", se= FALSE, color = "firebrick4", size = 1) +
  xlab(NULL) +
  ylab("Mean age") +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  # theme(plot.margin = margin(0, 0, 0, 0)) +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  coord_flip() +
  xlim(ymin, ymax) 


# Create the final plot
p_left <- plot_grid(p_dis_1, p_pcoa, nrow = 2, rel_heights = c(0.7,4), align = 'v')
p_right <- plot_grid(NULL, p_dis_2, nrow = 2, rel_heights = c(0.7,4))
p <- plot_grid(p_left, p_right, ncol = 2, rel_widths = c(4,0.7))

p # Figure 3C
```


#### Uniqueness

```{r fig.height=2.5, fig.width=2, message=FALSE, warning=FALSE}

min_non_zero <- function(x) {
  min_val <- min(x[x > 0], na.rm = TRUE)
  ifelse(is.infinite(min_val), NA, min_val)
}

# Apply the function across rows to get the minimal non-zero value for each row
uni <- bc %>%
  rowwise() %>%
  mutate(uniqueness = min_non_zero(c_across(-sample_id)))

# Select only the sample_id and uniqueness columns to create the new dataframe
uni <- select(uni, sample_id, uniqueness)

res <- get_mixed_effect(uni, 
                        sample_map %>% filter(!is.na(weight_from_vet_or_survey_lbs)), 
                        sample_id_column = 'sample_id', 
                        metadata_cols_to_analyze = c("age_at_stool_collection") , 
                        fdr_threshold = 0.05,
                        fixed_cols = FIXED,
                        random_cols = RANDOM)
if (length(res$plots) > 0){
  print(res$plots[[1]] + labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value"))  )
} else {
  print("No significant correlations to plot.")
}
  
res$plots[[1]] + # Figure 3D
  labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value")) + 
  labs(title = names(distances_dfs)[[i]])
  

```


Control for alpha diversity

```{r}

tmp_sample_map <- sample_map %>% 
  inner_join(shannon %>% select(sample_id, shannon_sp), by = 'sample_id')

min_non_zero <- function(x) {
  min_val <- min(x[x > 0], na.rm = TRUE)
  ifelse(is.infinite(min_val), NA, min_val)
}

for (i in 1:length(distances_dfs)){
  
  dist_df <- distances_dfs[[i]]
  # Apply the function across rows to get the minimal non-zero value for each row
  uni <- dist_df %>%
    rowwise() %>%
    mutate(uniqueness = min_non_zero(c_across(-sample_id)))
  
  # Select only the sample_id and uniqueness columns to create the new dataframe
  uni <- select(uni, sample_id, uniqueness)
  
  res <- get_mixed_effect(uni, 
                               tmp_sample_map, 
                               sample_id_column = 'sample_id', 
                               metadata_cols_to_analyze = c("age_at_stool_collection") , 
                               fdr_threshold = 0.05,
                               fixed_cols = c(FIXED, "shannon_sp"),
                               random_cols = RANDOM)
}

```

Healthiest dogs only

```{r fig.height=2.5, fig.width=2, message=FALSE, warning=FALSE}


min_non_zero <- function(x) {
  min_val <- min(x[x > 0], na.rm = TRUE)
  ifelse(is.infinite(min_val), NA, min_val)
}


# Apply the function across rows to get the minimal non-zero value for each row
uni <- bc %>%
  rowwise() %>%
  mutate(uniqueness = min_non_zero(c_across(-sample_id)))

# Select only the sample_id and uniqueness columns to create the new dataframe
uni <- select(uni, sample_id, uniqueness)

res <- get_mixed_effect(uni, 
                        sample_map %>% 
                          filter(!is.na(weight_from_vet_or_survey_lbs)) %>%
                          filter(hs_general_health <= 2), 
                        sample_id_column = 'sample_id', 
                        metadata_cols_to_analyze = c("age_at_stool_collection") , 
                        fdr_threshold = 0.05,
                        fixed_cols = FIXED,
                        random_cols = RANDOM)
if (length(res$plots) > 0){
  print(res$plots[[1]] + labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value"))  )
} else {
  print("No significant correlations to plot.")
}
  
```

Non-healthy dogs only

```{r fig.height=2.5, fig.width=2, message=FALSE, warning=FALSE}


min_non_zero <- function(x) {
  min_val <- min(x[x > 0], na.rm = TRUE)
  ifelse(is.infinite(min_val), NA, min_val)
}


# Apply the function across rows to get the minimal non-zero value for each row
uni <- bc %>%
  rowwise() %>%
  mutate(uniqueness = min_non_zero(c_across(-sample_id)))

# Select only the sample_id and uniqueness columns to create the new dataframe
uni <- select(uni, sample_id, uniqueness)

res <- get_mixed_effect(uni, 
                        sample_map %>% 
                          filter(!is.na(weight_from_vet_or_survey_lbs)) %>%
                          filter(hs_general_health >= 3), 
                        sample_id_column = 'sample_id', 
                        metadata_cols_to_analyze = c("age_at_stool_collection") , 
                        fdr_threshold = 0.05,
                        fixed_cols = FIXED,
                        random_cols = RANDOM)
if (length(res$plots) > 0){
  print(res$plots[[1]] + labs(subtitle = str_replace(res$plots[[1]]$label$subtitle, "FDR", "p-value"))  )
} else {
  print("No significant correlations to plot.")
}
  
```

Balanced life stage groups

```{r}

set.seed(42)  # for reproducibility

tmp_sample_map <- sample_map %>%
  filter(life_stage_at_stool_collection != "Senior") %>%
  filter(life_stage_at_stool_collection != "Puppy") %>%
  filter(hs_general_health < 3)

df_balanced <- tmp_sample_map %>%
  group_by(life_stage_at_stool_collection) %>%
  slice_sample(n = min(table(tmp_sample_map$life_stage_at_stool_collection))) %>%
  ungroup()

uni <- bc %>% select(sample_id, df_balanced$sample_id) %>%
  filter(sample_id %in% df_balanced$sample_id) %>%
  rowwise() %>%
  mutate(uniqueness = min_non_zero(c_across(-sample_id)))

# Select only the sample_id and uniqueness columns to create the new dataframe
uni <- select(uni, sample_id, uniqueness)

res_balanced <- get_mixed_effect2(uni, 
                                  df_balanced %>% filter(!is.na(weight_from_vet_or_survey_lbs)),# %>%
                                    # filter(hs_general_health < 2), 
                                  sample_id_column = 'sample_id', 
                                  metadata_cols_to_analyze = c("age_at_stool_collection") , 
                                  fdr_threshold = 1, #0.05,
                                  fixed_cols = FIXED,
                                  random_cols = RANDOM)

p <- res_balanced$plots$`uniqueness _ age_at_stool_collection` + labs(title = NULL, subtitle = NULL, y = "Uniqueness (Bray-Curtis)", x = "Age at stool collection")

p # Supplementary Figure S4 

```

# Section 2: specific species shifts - aging markers

## Continuous analysis

```{r}
res_ph <- run_continuous_maaslin(feat_table = ra_kraken_ph,
                                metadata =  sample_map %>% 
                                 filter(!is.na(weight_from_vet_or_survey_lbs)) %>%  
                                 mutate(breed_size_numeric = as.numeric(breed_size_numeric)), 
                                sample_id_column = 'sample_id', 
                                metadata_cols_to_analyze = c("age_at_stool_collection"), 
                                metadata_cols_fix_effects = c(FIXED),
                                metadata_cols_rand_effects = RANDOM,
                                fdr_threshold = 0.05,
                                output_dir = "da",
                                transformation_method = "LOG",
                                keep_maaslin_plots = FALSE,
                                category_name_map=NULL,
                                logscale_y = TRUE,
                                return_maaslin_plots = FALSE)

res_ph %>% write.csv(file.path(results_dir, "da_ph.csv"), row.names = FALSE)
```


```{r}
res_ge <- run_continuous_maaslin(feat_table = ra_kraken_ge,
                                metadata =  sample_map %>% 
                                 filter(!is.na(weight_from_vet_or_survey_lbs)) %>%  
                                 mutate(breed_size_numeric = as.numeric(breed_size_numeric)), 
                                sample_id_column = 'sample_id', 
                                metadata_cols_to_analyze = c("age_at_stool_collection"), 
                                metadata_cols_fix_effects = c(FIXED),
                                metadata_cols_rand_effects = RANDOM,
                                fdr_threshold = 0.05,
                                output_dir = "da",
                                transformation_method = "LOG",
                                keep_maaslin_plots = FALSE,
                                category_name_map=NULL,
                                logscale_y = TRUE,
                                return_maaslin_plots = FALSE)

res_ge %>% write.csv(file.path(results_dir, "da_ge.csv"), row.names = FALSE)
```


```{r fig.height=4, fig.width=5.5, message=FALSE, warning=FALSE}

res_sp <- run_continuous_maaslin(feat_table = ra_kraken_sp,
                                metadata =  sample_map %>% 
                                 filter(!is.na(weight_from_vet_or_survey_lbs)) %>%  
                                 mutate(breed_size_numeric = as.numeric(breed_size_numeric)), 
                                sample_id_column = 'sample_id', 
                                metadata_cols_to_analyze = c("age_at_stool_collection"), 
                                metadata_cols_fix_effects = c(FIXED),
                                metadata_cols_rand_effects = RANDOM,
                                fdr_threshold = 0.05,
                                output_dir = "da",
                                transformation_method = "LOG",
                                keep_maaslin_plots = FALSE,
                                category_name_map=NULL,
                                logscale_y = TRUE,
                                return_maaslin_plots = FALSE)

res_sp %>% write.csv(file.path(results_dir, "da_sp.csv"), row.names = FALSE)

# Load the GTDB taxonomy to get the genus and phylum per each species
gtdb <- samples_metadata <- read_tsv(
  file.path(metadata_dir, "gtdb", "bac120_taxonomy_r207.tsv"),show_col_types = FALSE, col_names=FALSE) %>%
  rename(taxonomy = X2) %>%   
  separate(taxonomy, into = c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), sep = ";") %>%
  mutate(
    Domain = sub("d__", "", Domain),
    Phylum = sub("p__", "", Phylum),
    Class = sub("c__", "", Class),
    Order = sub("o__", "", Order),
    Family = sub("f__", "", Family),
    Genus = sub("g__", "", Genus),
    Species = sub("s__", "", Species)
  )

# Plot species with significant association with age
p <- res_sp$all_results %>%
  filter(value == "age_at_stool_collection") %>%
  mutate(Genus = str_extract(feature, "^[^.]+")) %>%
  left_join(gtdb %>% select(Genus, Phylum) %>% distinct(), by = "Genus") %>%
    mutate(Genus = case_when(
    Genus %in% c("Prevotella", "Holdemanella") ~ Genus,
    TRUE ~ "other"
  )) %>%
  ggplot(aes(x = coef, y = -log(FDR), fill=Genus, color=Genus)) +
  geom_point( alpha = 0.7, shape = 21, size = 2.5) +
  geom_hline(yintercept = -log(0.05), linetype = "dashed", color = "red") +  
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + 
  labs(x = "Regression coefficient", y = "-log(FDR)") +  # Axis labels
  # Color specific genus
  scale_fill_manual(values = c("Prevotella" = "#f7b267",  "Holdemanella" = "#c9184a", "other" = "grey80"), 
                    breaks = c("Prevotella", "Holdemanella", "C")) +
  scale_color_manual(values = c("Prevotella" = "grey20",  "Holdemanella" = "grey20", "other" = "grey20"),
                     breaks = c("Prevotella", "Holdemanella", "C")) +
  theme_classic()

p # Figure 3E
```


## Categorical analysis

```{r}
get_differential_abundance <- function(feat_table, 
                                       disct_metadata, 
                                       sample_id_column = 'sample_id', 
                                       metadata_cols_to_analyze = NULL,
                                       metadata_cols_fix_effects = NULL,
                                       metadata_cols_rand_effects = NULL,
                                       fdr_threshold = 0.1,
                                       output_dir = "MaAsLin2_output",
                                       return_maaslin_plots = FALSE,
                                       keep_maaslin_plots = FALSE,
                                       normalization_method = "NONE",
                                       transformation_method = "LOG",
                                       analysis_method = "LM",
                                       category_name_map = NULL,
                                       skip_fdr_correction = FALSE,
                                       logscale_y_for_plot = FALSE,
                                       quiet = FALSE,
                                       ord=NULL) {
  # Required libraries loading
  require(Maaslin2)
  require(ggplot2)
  require(ggsignif)
  require(grid)
  require(png)
  require(dplyr)
  require(ggpubr)
  require(purrr)
  
  # Data verifications
  
  # In case maaslin plots required, change keep_maaslin_plots to TRUE.
  if (return_maaslin_plots & !(keep_maaslin_plots)){
    message("Can not return maaslin plots if keep_maaslin_plots = FALSE. keep_maaslin_plots was changed to TRUE.\n")
    keep_maaslin_plots = TRUE
  }
  
  # If no metadata_cols_to_analyze provided, run the MaAsLin2 on all the metadata columns.
  if (is.null(metadata_cols_to_analyze))
    metadata_cols_to_analyze = colnames(disct_metadata)
  
  # Check if metadata columns are discrete (1, 0 and NA).
  for (col in metadata_cols_to_analyze){ 
    if (length(unique(disct_metadata[,col])) > 3)
      warning(paste0("Too many values in disct_metadata in the ", col, " column"))
  }
  
  # Check if normalization method fits the data (if only one column: normalization method cannot be 'TSS')
  if ((ncol(feat_table) ==  2) & (normalization_method == "TSS")){
    stop("TSS normlaization is not allowed in case there is only one feature in feat_table. 
         Try again with normalization_method = 'NONE'.")
  }
  
  # Create output directory (will be deleted if not required).
  if (!dir.exists(output_dir))
    dir.create(output_dir)
  
  # Convert sample_id column to rownames for MaAsLin2 run.
  disct_metadata <- disct_metadata %>% remove_rownames %>% column_to_rownames(var=sample_id_column)
  feat_table <- feat_table %>% remove_rownames %>% column_to_rownames(var=sample_id_column)
  
  # Remove samples not in both tables, and order samples identically
  samples_to_include <- intersect(row.names(disct_metadata), row.names(feat_table))
  disct_metadata <- disct_metadata[samples_to_include,]
  feat_table <- feat_table[samples_to_include,,drop=FALSE]
  
  # TODO: warn if some samples were not in both tables
  
  # Check that all the metadata_cols_rand_effects are not in metadata_cols_to_analyze.
  for (rand_col in metadata_cols_rand_effects){
    if (rand_col %in% metadata_cols_to_analyze)
      stop(paste0("Shared column between metadata_cols_rand_effects and metadata_cols_to_analyze: ", 
                  rand_col, "\nTry again with different metadata_cols_rand_effects!\n"))
  }
  
  # Go over fixed effects and perform one-hot encoding if needed
  fixed_cols_final_names <- c()
  for (fixed_col in metadata_cols_fix_effects) {

    # Validate that metadata_cols_fix_effects are not in metadata_cols_to_analyze.
    if (fixed_col %in% metadata_cols_to_analyze)
      stop(paste0("Shared column between metadata_cols_fix_effects and metadata_cols_to_analyze: ", 
                  fixed_col, "\nTry again with different metadata_cols_fix_effects!\n"))
    
    # Check if fixed column is numeric with less than 5 unique values or categorical. Perform one-hot encoding in these cases.
    if (((is.numeric(disct_metadata[, fixed_col])) & 
         (n_distinct(disct_metadata[, fixed_col]) < 5)) | 
        (!is.numeric(disct_metadata[, fixed_col]))) {
      
      # Print a message if there are more than 5 categories in the metadata_cols_fix_effects because it might affect the FDR correction.
      if (n_distinct(disct_metadata[[fixed_col]]) > 5)
        message("Too many unique values in ", fixed_col, " column. It might affect the FDR correction and omit some significant results!")
      
      # Perform one-hot encoding for columns if they have more then 2 categories.
      if (n_distinct(disct_metadata[[fixed_col]]) > 2){
        message("Performing one-hot encoding on ", fixed_col)
        fixed_cols_final_names <- c(fixed_cols_final_names, paste(fixed_col, unique(disct_metadata[[fixed_col]]), sep = '_'))
        tmp_metadata_rownames <- rownames(disct_metadata)
        disct_metadata <- disct_metadata %>% mutate(value = 1)  %>% spread(fixed_col, value,  fill = 0, sep = '_')
        rownames(disct_metadata) <- tmp_metadata_rownames
        
      } else { # Categorical with 2 values or less. One-hot is not needed 
        fixed_cols_final_names <- c(fixed_cols_final_names, fixed_col)
      }
      
    } else { # Numeric case, one-hot is not needed  
      fixed_cols_final_names <- c(fixed_cols_final_names, fixed_col)
    }
  }
  
  # Rum MaAsLin2
  maaslin2_signif_results <- data.frame() # Initialize
  maaslin2_all_results <- data.frame() # Initialize
  
  for (col in metadata_cols_to_analyze) {
    if (!quiet) message('Working on column: ', col)
    
    # For multi-category variables, we run differential abundance separately for each pair of categories
    for (pair in combn(na.omit(unique(disct_metadata[[col]])), 2, simplify = FALSE)) {
      
      # Keep only the two selected values (also drop missing values)
      tmp_metadata <- disct_metadata[(!is.na(disct_metadata[[col]])) & (disct_metadata[[col]] %in% pair), ]
      
      # Get the final list of samples for which DA will be calculated
      samples_to_include <- intersect(row.names(tmp_metadata), row.names(feat_table))
      tmp_metadata <- tmp_metadata[samples_to_include,,drop=F]
      tmp_feat_table <- feat_table[samples_to_include,,drop=F]
      
      # Remove constant columns if exist
      tmp_metadata <- tmp_metadata[, apply(tmp_metadata, 2, n_distinct) > 1]
      
      # Fix column names if needed
      colnames(tmp_metadata) <- make.names(colnames(tmp_metadata))
      if (!is.null(metadata_cols_rand_effects))
        metadata_cols_rand_effects <- make.names(metadata_cols_rand_effects)
      
      invisible(capture.output(tmp_results <- Maaslin2(
        input_data = tmp_feat_table, 
        input_metadata = tmp_metadata, 
        output = paste(output_dir, "/", col, sep=""),
        fixed_effects = make.names(c(col, fixed_cols_final_names)),
        random_effects = metadata_cols_rand_effects,
        normalization = normalization_method,
        transform = transformation_method,
        analysis_method = analysis_method,
        plot_scatter = keep_maaslin_plots,
        max_significance = fdr_threshold,
        plot_heatmap = FALSE
      )))
      
      tmp_results$results$pair_a <- as.character(pair[[1]])
      tmp_results$results$pair_b <- as.character(pair[[2]])
      
      # Save significant results only
      maaslin2_signif_results <- bind_rows(
        maaslin2_signif_results,
        tmp_results$results %>% filter(qval <= fdr_threshold)
      )
      
      # Save all results
      maaslin2_all_results <- bind_rows(
        maaslin2_all_results,
        tmp_results$results
      )
    }
  }
  
  # New FDR 
  if (skip_fdr_correction){
    maaslin2_all_results$FDR <- maaslin2_all_results$pval
  } else {
    maaslin2_all_results$FDR <- p.adjust(maaslin2_all_results$pval, 'fdr')
  }
  maaslin2_signif_results <- maaslin2_all_results[maaslin2_all_results$FDR < fdr_threshold,]
  
  # Return significant results only for column in metadata_cols_to_analyze (and not in metadata_cols_fix_effects).
  maaslin2_signif_results <- maaslin2_signif_results[maaslin2_signif_results$metadata %in% metadata_cols_to_analyze, ]
  
  # Generate plots to plots list
  plots <- list()
  
  if (nrow(maaslin2_signif_results) > 0) {
    if (return_maaslin_plots) {
      png_list <- fs::dir_ls(path = output_dir, recurse = TRUE, type = "file", glob = "*.png")
      for (png_dir in png_list){
        print(png_dir)
        plots[[png_dir]] <- rasterGrob(readPNG(png_dir))
      }
    } else {
      
      # Fix feature names to match how maaslin outputs them
      feat_table2 <- feat_table
      names(feat_table2) <- make.names(names(feat_table2))
      
      
      for (col in unique(maaslin2_signif_results$metadata)){
        for (feat in unique(maaslin2_signif_results$feature)){
          
          unique_signif_results <- maaslin2_signif_results[(maaslin2_signif_results$metadata == col) & (maaslin2_signif_results$feature == feat), ]
          if (dim(unique_signif_results)[1] == 0)
            next
            

          tmp_metadata <- disct_metadata %>% tidyr::drop_na(col)
          samples_to_include <- intersect(row.names(tmp_metadata), row.names(feat_table2))
          tmp_metadata <- tmp_metadata[samples_to_include,,drop=FALSE]
          feat_table2 <- feat_table[samples_to_include,,drop=FALSE]
          
          tmp <- data.frame(
            col_vals = tmp_metadata[[col]],
            feat_vals = feat_table2[[feat]]
          )
          
          if (!is.null(category_name_map)) 
            tmp$col_vals <- factor(category_name_map[as.character(tmp$col_vals)],
                                   levels = unname(category_name_map))
          
          y_m <- max(tmp$feat_vals) 
          tmp$col_vals<- as.character(tmp$col_vals)
          
          # Add number of samples in each group to the x-axis ticks.
          xlabs <- list()
          for (val in as.character(unique(tmp$col_vals)))
            xlabs[[val]] = paste0(val,"\n(", dim(tmp[as.character(tmp$col_vals) == val,])[1], ")")
          
          # Sort the tmp dataset based on the specified order
          if (!is_null(ord)){     
            tmp$col_vals <- factor(tmp$col_vals, levels = ord)
            tmp <- tmp[order(tmp$col_vals), ]
          }
          
          # p <- ggplot(tmp, aes(x = col_vals, group = col_vals, y = feat_vals)) +
          #   geom_boxplot(aes(fill = col_vals), color = 'black', alpha = 0.4, outlier.shape = NA) +                      geom_jitter(height = 0, width = 0.1, alpha = 0.2, color = 'black', size = 2) +
          #   theme_classic() +
          #   xlab(col) +
          #   ylab(feat) +
          #   # labs(subtitle = 'MaAsLin2') +
          #   theme(plot.subtitle = element_text(hjust = 0.5)) +
          #   scale_x_discrete(labels=xlabs) +
          #   scale_fill_manual(values = list("Puppy" = "#CA0020", "Young Adult" = "#F4A582",  
          #                                   "Mature Adult" = "#92C5DE", "Senior" = "#0571B0")) +
          #   theme(legend.position = "none")
          
          p <- ggplot(tmp, aes(x = col_vals, group = col_vals, y = feat_vals)) +
              geom_boxplot(aes(fill = col_vals), color = 'black', alpha = 0.4, outlier.shape = NA) +
              geom_jitter(height = 0, width = 0.1, alpha = 0.2, color = 'black', size = 2) +
              theme_classic() +
              xlab(gsub("_", " ", tools::toTitleCase(col))) +
              ylab(feat) +
              theme(plot.subtitle = element_text(hjust = 0.5)) +
              scale_x_discrete(labels = xlabs) +
              scale_fill_manual(values = c(
                "Puppy" = "#CA0020",
                "Young Adult" = "#F4A582",
                "Mature Adult" = "#92C5DE",
                "Senior" = "#0571B0"
              )) +
              theme(legend.position = "none")
          
          
          unique_signif_results$group1 <- unique_signif_results$pair_a
          unique_signif_results$group2 <- unique_signif_results$pair_b
          
          makeStars <- function(x){
            stars <- c("****", "***", "**", "*")
            vec <- c(0, fdr_threshold/1000, fdr_threshold/100, fdr_threshold/10 ,fdr_threshold)
            i <- findInterval(x, vec)
            stars[i]
          }
          
          unique_signif_results$FDR <- makeStars(unique_signif_results$FDR)
          
          p <- p + stat_pvalue_manual(unique_signif_results,
                                      y.position = max(tmp$feat_vals),
                                      step.increase = 0.1 ,
                                      label = "FDR")

          # Log scale y axis if needed
          if (logscale_y_for_plot) {
            p <- p + yscale("log10", .format = TRUE)
          }
          plots[[paste(col, feat, collapse = ';')]] <- p
        }
      }
    } 
  }
  
  if (! keep_maaslin_plots){
    unlink(output_dir, recursive=TRUE)
  }
  
  
  return(list(sig_results = maaslin2_signif_results, all_results = maaslin2_all_results, plots = plots))
}
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(gridExtra)

# --------------------------------------------------
# Global settings
# --------------------------------------------------
sample_id_col <- "sample_id"
life_stage_levels <- c("Puppy", "Young Adult", "Mature Adult", "Senior")
main_metadata_col <- "life_stage_at_stool_collection"

# --------------------------------------------------
# Helper function
# --------------------------------------------------
run_da_summary <- function(feat_table, sample_map, output_dir_name) {
  
  # ----------------------------------------------
  # 1. Run differential abundance
  # ----------------------------------------------
  res <- get_differential_abundance(
    feat_table = feat_table,
    disct_metadata = sample_map %>%
      filter(!is.na(weight_from_vet_or_survey_lbs)) %>%
      mutate(breed_size_numeric = as.numeric(breed_size_numeric)),
    sample_id_column = sample_id_col,
    metadata_cols_to_analyze = c(main_metadata_col),
    metadata_cols_fix_effects = FIXED,
    metadata_cols_rand_effects = RANDOM,
    fdr_threshold = 0.05,
    output_dir = output_dir_name,
    transformation_method = "LOG",
    keep_maaslin_plots = FALSE,
    category_name_map = NULL,
    logscale_y = TRUE,
    return_maaslin_plots = FALSE,
    ord = life_stage_levels
  )
  
  # ----------------------------------------------
  # 2. Merge feature table with metadata
  #    and remove missing life-stage values
  # ----------------------------------------------
  df <- feat_table %>%
    inner_join(
      sample_map %>% select(sample_id, all_of(main_metadata_col)),
      by = "sample_id"
    ) %>%
    filter(!is.na(.data[[main_metadata_col]])) %>%
    mutate(
      !!main_metadata_col := factor(
        .data[[main_metadata_col]],
        levels = life_stage_levels
      )
    ) %>%
    droplevels()
  
  # ----------------------------------------------
  # 3. Convert to long format
  # ----------------------------------------------
  df_long <- df %>%
    pivot_longer(
      cols = -c(all_of(sample_id_col), all_of(main_metadata_col)),
      names_to = "taxon",
      values_to = "abundance"
    )
  
  # ----------------------------------------------
  # 4. Summary statistics by taxon and life stage
  # ----------------------------------------------
  summary_stats <- df_long %>%
    group_by(.data[[main_metadata_col]], taxon) %>%
    summarise(
      median_abundance = median(abundance, na.rm = TRUE),
      q25 = quantile(abundance, 0.25, na.rm = TRUE),
      q75 = quantile(abundance, 0.75, na.rm = TRUE),
      prevalence = mean(abundance > 0, na.rm = TRUE),
      .groups = "drop"
    )
  
  summary_stats_wide <- summary_stats %>%
    pivot_wider(
      id_cols = taxon,
      names_from = all_of(main_metadata_col),
      values_from = c(median_abundance, q25, q75, prevalence),
      names_sep = "__"
    )
  
  # ----------------------------------------------
  # 5. MaAsLin results in wide format
  # ----------------------------------------------
  maaslin_wide <- res$all_results %>%
    filter(metadata == main_metadata_col) %>%
    mutate(comparison = paste(pair_a, pair_b, sep = "_vs_")) %>%
    select(feature, comparison, coef, stderr, pval, FDR) %>%
    distinct() %>%
    pivot_wider(
      id_cols = feature,
      names_from = comparison,
      values_from = c(coef, stderr, pval, FDR),
      names_sep = "__"
    ) %>%
    rename(taxon = feature)
  
  # ----------------------------------------------
  # 6. Final merged table
  # ----------------------------------------------
  final_table <- summary_stats_wide %>%
    left_join(maaslin_wide, by = "taxon")
  
  # ----------------------------------------------
  # 7. Order taxa by mean abundance
  # ----------------------------------------------
  top_taxa <- df_long %>%
    group_by(taxon) %>%
    summarise(
      mean_abundance = mean(abundance, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(mean_abundance)) %>%
    pull(taxon)
  
  print(top_taxa[1:min(20, length(top_taxa))])
  
  # ----------------------------------------------
  # 8. Select first 16 existing plots by top_taxa order
  # ----------------------------------------------
  plot_names <- names(res$plots)
  
  selected_plot_names <- c()
  
  for (tx in top_taxa) {
    this_plot_name <- paste(main_metadata_col, tx)
    
    if (this_plot_name %in% plot_names) {
      selected_plot_names <- c(selected_plot_names, this_plot_name)
    }
    
    if (length(selected_plot_names) == 16) {
      break
    }
  }
  
  plot_list_16 <- res$plots[selected_plot_names]
  
  plot_grid <- arrangeGrob(
    grobs = plot_list_16,
    ncol = 4,
    nrow = 4
  )
  
  # ----------------------------------------------
  # 9. Return all outputs
  # ----------------------------------------------
  list(
    res = res,
    df = df,
    df_long = df_long,
    summary_stats = summary_stats,
    summary_stats_wide = summary_stats_wide,
    maaslin_wide = maaslin_wide,
    final_table = final_table,
    top_taxa = top_taxa,
    selected_plot_names = selected_plot_names,
    plot_list_16 = plot_list_16,
    plot_grid = plot_grid
  )
}
```


```{r}
# --------------------------------------------------
# Run for species
# --------------------------------------------------
out_species <- run_da_summary(
  feat_table = ra_kraken_sp,# %>% select(1:3),
  sample_map = sample_map,
  output_dir_name = "da_species"
)

# --------------------------------------------------
# Run for genus
# --------------------------------------------------
out_genus <- run_da_summary(
  feat_table = ra_kraken_ge,# %>% select(1:30),
  sample_map = sample_map,
  output_dir_name = "da_genus"
)

# --------------------------------------------------
# Run for phylum
# --------------------------------------------------
out_phylum <- run_da_summary(
  feat_table = ra_kraken_ph,# %>% select(1:30),
  sample_map = sample_map,
  output_dir_name = "da_phylum"
)

# --------------------------------------------------
# Run for KEGG modules
# --------------------------------------------------
out_kegg_modules <- run_da_summary(
  feat_table = kegg_modules %>% select(-unknown),# %>% select(1:3),
  sample_map = sample_map,
  output_dir_name = "da_kegg_modules"
)

# --------------------------------------------------
# Run for KEGG pathways
# --------------------------------------------------
out_kegg_pathway <- run_da_summary(
  feat_table = kegg_pwys %>% select(-unknown),# %>% select(1:3),
  sample_map = sample_map,
  output_dir_name = "da_kegg_pathway"
)

```


```{r}
grid::grid.newpage()
grid::grid.draw(out_species$plot_grid) # Supplementary Figure S5

grid::grid.newpage()
grid::grid.draw(out_genus$plot_grid) # Supplementary Figure S6
```


```{r}
modules_anno <- read.delim("modules_from_kegg.csv", 
                      sep = ",") %>%
  rename(feature = entry, kegg_description = module_name)

ko_anno <- read.delim("pathways_from_kegg.list", 
                      sep = "\t", skip = 1) %>%
  rownames_to_column(var = "feature") %>%
  rename(kegg_description = X..Global.and.overview.maps) %>%
  # Add prefix of "ko" at the feature name
  mutate(feature = paste0("ko", feature))

kegg_anno <- bind_rows(ko_anno, modules_anno)

supp_table <- bind_rows(
  out_species$final_table %>% mutate(feature_type = "species"),
  out_genus$final_table %>% mutate(feature_type = "genus"),
  out_phylum$final_table %>% mutate(feature_type = "phylum"),
  out_kegg_modules$final_table %>% mutate(feature_type = "kegg_module"),
  out_kegg_pathway$final_table %>% mutate(feature_type = "kegg_pathway")
) %>%
  relocate(feature_type, .before = taxon) %>%
  rename(feature = taxon) %>% 
  left_join(kegg_anno, by = "feature") %>%
  relocate(kegg_description, .before = median_abundance__Puppy) 

supp_table %>%
  write.csv("supplementary_table_s12.csv", row.names = FALSE)

```


## Compare to human aging markers (Lifelines)

```{r}

# Read Lifelines kraken table (genus level) and reorganize it

ll_kraken_ge <- read_tsv2(file.path(ll_feat_table_dir, "kraken_genus_level_taxonomy.tsv")) %>%
  convert_to_relab()

prevalence_cutoff = 0.2 
avg_abundance_cutoff = 0.0001

ll_kraken_ge <- remove_rare_features(ll_kraken_ge, prevalence_cutoff=prevalence_cutoff, avg_abundance_cutoff=avg_abundance_cutoff)
```

```{r, warning=FALSE}

# Read Lifelines metadata and reorganize it

ll_metadata <- read_csv(file.path(ll_feat_table_dir, "samples.csv"), show_col_types = FALSE) %>%
  mutate(age = str_extract(title, "(?<=Age:)\\d+")) %>%
  mutate(alias = str_remove(alias, "LL$"))

bam_files <- read_csv(file.path(ll_feat_table_dir, "bam_files.csv"), show_col_types = FALSE) %>%
  # Get new alias col that is bam col without the .bam suffix
  mutate(alias = str_remove(bam, "\\.bam$")) %>%
  # remove the LL sufix from the alias col if there is 
  mutate(alias = str_remove(alias, "LL$"))

ll_metadata <- ll_metadata %>%
  inner_join(bam_files, by = "alias") %>%
  rename(sample_id = subdir) %>%
  # select(sample_id, age) %>%
  mutate(age = as.numeric(age)) 

```

```{r}

source('continuous_metadata_analysis.R')

ll_res <- run_continuous_maaslin(feat_table = ll_kraken_ge,
                                  metadata =  ll_metadata, 
                                  sample_id_column = 'sample_id', 
                                  metadata_cols_to_analyze = c("age"), 
                                  # metadata_cols_fix_effects = FIXED,
                                  # metadata_cols_rand_effects = RANDOM,
                                  fdr_threshold = 0.05,
                                  output_dir = "da",
                                  transformation_method = "LOG",
                                  keep_maaslin_plots = FALSE,
                                  category_name_map=NULL,
                                  logscale_y = TRUE,
                                  return_maaslin_plots = FALSE)
```

```{r, fig.height=8, fig.width=8, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(ggrepel)
library(cowplot)

df <- res_ge$all_results %>% 
  filter(metadata == "age_at_stool_collection") %>%
  rename(
    dogs_fdr   = FDR,
    dog_coef   = coef,
    dog_stderr = stderr,
    dog_pvalue = pval,
    dog_value  = value
  ) %>%
  select(feature, dog_coef, dogs_fdr, dog_stderr, dog_pvalue) %>%
  inner_join(
    ll_res$all_results %>%
      rename(
        human_fdr    = FDR,
        human_coef   = coef,
        human_stderr = stderr,
        human_pvalue = pval,
        human_value  = value
      ),
    by = "feature"
  ) %>%
  mutate(
    sig_dog   = dogs_fdr  < 0.1,
    sig_human = human_fdr < 0.1,
    sig_both  = sig_dog & sig_human,
    status = case_when(
      sig_both               ~ "both",
      sig_dog & !sig_human   ~ "dog_only",
      sig_human & !sig_dog   ~ "human_only",
      TRUE                   ~ "none"
    ),
    status = factor(status, levels = c("none", "dog_only", "human_only", "both"))
  )

cor(df$dog_coef, df$human_coef, method = "spearman")

# --------------------
# Main plot (NO legend)
# --------------------
p_main <- ggplot(df, aes(x = human_coef, y = dog_coef)) +
  # NONE: gray dot only (no outline, no inner dot)
  geom_point(
    data = df %>% filter(status == "none"),
    shape = 21, size = 2.7,
    fill = "grey90", color = "grey85",
    alpha = 0.8
  ) +
  # DOG ONLY: gray dot + black border
  geom_point(
    data = df %>% filter(status == "dog_only"),
    shape = 21, size = 2.7,
    fill = "grey90", color = "black", stroke = 0.9,
    alpha = 0.9
  ) +
  # HUMAN ONLY: gray background + black inner dot
  geom_point(
    data = df %>% filter(status == "human_only"),
    shape = 21, size = 2.7,
    fill = "grey90", color = "grey85",
    alpha = 0.9
  ) +
  geom_point(
    data = df %>% filter(status == "human_only"),
    shape = 16, size = 1.05,
    color = "black",
    alpha = 0.95
  ) +
  # BOTH: gray background + black border + black inner dot
  geom_point(
    data = df %>% filter(status == "both"),
    shape = 21, size = 2.7,
    fill = "grey90", color = "black", stroke = 0.9,
    alpha = 0.95
  ) +
  geom_point(
    data = df %>% filter(status == "both"),
    shape = 16, size = 1.05,
    color = "black",
    alpha = 0.98
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey70") +
  coord_cartesian(ylim = c(-1.2, 1.2)) +
  labs(
    x = "Human age coefficient",
    y = "Dog age coefficient"
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")

# --------------------
# Legend panel (manual, matches exactly)
# --------------------
# Legend panel (tight + looks like the plot)
legend_dat <- tibble::tibble(
  status = factor(
    c("none", "dog_only", "human_only", "both"),
    levels = c("none", "dog_only", "human_only", "both")
  ),
  label = c(
    "Not significant",
    "Significant in dogs only",
    "Significant in humans only",
    "Significant in dogs and humans"
  )
)

p_leg <- ggplot(legend_dat, aes(y = status)) +
  # base gray dot for all
  geom_point(
    aes(x = 1),
    shape = 21, size = 3.0,
    fill = "grey85", color = "grey85",
    alpha = 0.95

  ) +
  # border for dog_only + both
  geom_point(
    data = legend_dat %>% filter(status %in% c("dog_only", "both")),
    aes(x = 1),
    shape = 21, size = 3.0,
    fill = "grey85", color = "black", stroke = 0.9,
    alpha = 0.95

  ) +
  # inner dot for human_only + both
  geom_point(
    data = legend_dat %>% filter(status %in% c("human_only", "both")),
    aes(x = 1),
    shape = 16, size = 1.2,
    color = "black"
  ) +
  geom_text(
    aes(x = 1.25, label = label),
    hjust = 0, size = 3.4
  ) +
  scale_y_discrete(drop = FALSE) +
  scale_x_continuous(limits = c(0.9, 3.6), expand = c(0, 0)) +
  theme_void(base_size = 12) +
  theme(
    plot.margin = margin(0, 0, 0, 0),
    panel.spacing = grid::unit(0, "pt")
  )


# Combine plot + legend panel
p <- cowplot::plot_grid(p_main, p_leg, ncol = 2, rel_widths = c(1, 0.45))

p # Supplementary Figure S7

```


# Section 3: predict age based on microbiome features

```{r}
get_predictability_regression_for_age_pred <- function(feat_table, 
                                                       metadata, 
                                                       sample_id_column, 
                                                       vars_to_analyze, 
                                                       ml_method = 'rf',
                                                       fdr_threshold = 0.1,
                                                       corr_threshold = 0.3,
                                                       corr_method = "pearson",
                                                       n_ml_repeats = 10,
                                                       n_ml_folds = 5,
                                                       feat_importance = 'none',
                                                       rand_seed = 1111,
                                                       quiet = FALSE){
  # Required libraries loading
  require(rsample)
  require(dplyr)
  require(glmnet)
  require(ranger)
  require(metap)
  require(tibble)
  
  # Argument validations 
  if (is.null(vars_to_analyze))
    stop("Invalid 'vars_to_analyze' argument. Cannot be NULL")
  if (!all(vars_to_analyze %in% names(metadata)))
    stop("Invalid 'vars_to_analyze' argument. Some of the variables provided are missing from the metadata table")
  if (!ml_method %in% c('rf'))
    stop("Invalid 'ml_method' argument. Only 'rf' supported.")
  if (!corr_method %in% c('pearson', 'spearman'))
    stop("Invalid 'corr_method' argument. Only 'pearson' and 'spearman' supported.")
  
  # Metric helper functions
  rmse_fun <- function(obs, pred) {
    sqrt(mean((obs - pred)^2, na.rm = TRUE))
  }
  
  mae_fun <- function(obs, pred) {
    mean(abs(obs - pred), na.rm = TRUE)
  }
  
  r2_fun <- function(obs, pred) {
    obs <- as.numeric(obs)
    pred <- as.numeric(pred)
    ss_res <- sum((obs - pred)^2, na.rm = TRUE)
    ss_tot <- sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
    if (is.na(ss_tot) || ss_tot == 0) return(NA_real_)
    1 - ss_res / ss_tot
  }
  
  sumlog2 <- function(ps) {
    ps <- ps[!is.na(ps)]
    if (length(ps) == 0) return(NA_real_)
    metap::sumlog(ps)$p
  }
  
  # Main calculations
  n_samples <- nrow(feat_table)
  
  # Generate folds for cross-validation
  set.seed(rand_seed)
  folds <- vfold_cv(
    data.frame(sample_num = 1:n_samples),
    v = n_ml_folds,
    repeats = n_ml_repeats
  )
  
  # Collect evaluation stats here
  cv_metrics <- data.frame()
  cv_feat_imp <- data.frame()
  fold_preds_all <- list()
  
  # Fix names of features (causes errors in some models)
  names(feat_table) <- make.names(names(feat_table))
  
  # Iterate over variables
  for (measure in vars_to_analyze) {
    if (!quiet) message("Training model for variable: ", measure)
    
    # Prepare temporary table for training (features + label only)
    tmp <- feat_table %>% 
      left_join(
        metadata[, c(sample_id_column, measure)] %>% 
          rename(Label = 2), 
        by = sample_id_column
      )
    
    pred_per_fold_list <- list()
    
    # Cross validation loop
    for (i in 1:nrow(folds)) {
      fold_id <- paste(folds$splits[[i]]$id, collapse = " ")
      
      # Get train/test samples of this fold
      train_samples <- folds$splits[[i]]$in_id
      train_data <- tmp[train_samples, ] %>% filter(!is.na(Label))
      test_data  <- tmp[-train_samples, ] %>% filter(!is.na(Label))
      
      test_data_with_sample_id <- test_data
      
      train_data <- train_data %>%
        select(-any_of(c(sample_id_column)))
      test_data <- test_data %>%
        select(-any_of(c(sample_id_column)))
      
      # Subset training and testing data
      train_data_boruta <- train_data
      test_data_boruta  <- test_data
      
      # Train RF
      rf_model <- ranger(
        Label ~ ., 
        data = train_data_boruta, 
        importance = ifelse(feat_importance == 'altmann', 'permutation', 'none'),
        seed = rand_seed
      )
      
      if (feat_importance != 'none') {
        feat_imp <- importance_pvalues(
          rf_model, 
          method = "altmann", 
          formula = Label ~ ., 
          data = train_data_boruta, 
          seed = rand_seed
        )
      }
      
      # Make prediction on held out samples
      preds <- predict(rf_model, test_data_boruta %>% select(-Label))$predictions
      
      # Store fold predictions
      tmp_eval <- data.frame(
        measure = measure,
        fold_id = fold_id,
        preds = preds,
        true_label = test_data_boruta$Label,
        sample_id = test_data_with_sample_id[[sample_id_column]]
      )
      
      pred_per_fold_list[[i]] <- tmp_eval
      fold_preds_all[[paste0(measure, "__", i)]] <- tmp_eval
      
      # Correlation test only if enough variance / observations
      cor_val <- NA_real_
      cor_pval <- NA_real_
      
      if (nrow(tmp_eval) >= 3 &&
          stats::sd(tmp_eval$preds, na.rm = TRUE) > 0 &&
          stats::sd(tmp_eval$true_label, na.rm = TRUE) > 0) {
        
        cor_val <- suppressWarnings(
          cor(tmp_eval$preds, tmp_eval$true_label, method = corr_method, use = "complete.obs")
        )
        
        cor_pval <- suppressWarnings(
          cor.test(
            tmp_eval$preds,
            tmp_eval$true_label,
            alternative = 'greater',
            method = corr_method,
            exact = FALSE
          )$p.value
        )
      }
      
      # Collect per-fold metrics
      cv_metrics <- bind_rows(
        cv_metrics,
        data.frame(
          measure = measure,
          fold_id = fold_id,
          test_cor = cor_val,
          test_cor_p = cor_pval,
          rmse = rmse_fun(tmp_eval$true_label, tmp_eval$preds),
          mae = mae_fun(tmp_eval$true_label, tmp_eval$preds),
          r2 = r2_fun(tmp_eval$true_label, tmp_eval$preds)
        )
      )
      
      # Collect feature importance
      if (feat_importance != 'none') {
        cv_feat_imp <- bind_rows(
          cv_feat_imp,
          as.data.frame(feat_imp) %>% 
            tibble::rownames_to_column('feature') %>%
            mutate(measure = measure, fold_id = fold_id)
        )
      }
    }
  }
  
  # Summarise over folds
  summary_correlations <- cv_metrics %>%
    group_by(measure) %>%
    summarise(
      n_folds = n(),
      mean_cor = mean(test_cor, na.rm = TRUE),
      sd_cor = sd(test_cor, na.rm = TRUE),
      mean_rmse = mean(rmse, na.rm = TRUE),
      sd_rmse = sd(rmse, na.rm = TRUE),
      mean_mae = mean(mae, na.rm = TRUE),
      sd_mae = sd(mae, na.rm = TRUE),
      mean_r2 = mean(r2, na.rm = TRUE),
      sd_r2 = sd(r2, na.rm = TRUE),
      p = sumlog2(test_cor_p),
      .groups = "drop"
    ) %>%
    mutate(
      fdr = p.adjust(p, method = "fdr"),
      Significant = fdr < fdr_threshold & mean_cor > corr_threshold
    )
  
  # Summarize feature importance (if requested by user)
  summary_feat_importance <- data.frame()
  if (feat_importance != 'none') {
    summary_feat_importance <- cv_feat_imp %>%
      group_by(measure, feature) %>%
      summarise(
        n_folds = n(),
        mean_feat_importance = mean(importance, na.rm = TRUE),
        sd_feat_importance = sd(importance, na.rm = TRUE),
        p = sumlog2(pvalue),
        .groups = "drop"
      ) %>%
      mutate(fdr = p.adjust(p, method = "fdr"))
  }
  
  if (!any(summary_correlations$Significant, na.rm = TRUE)) {
    message('No significantly well-predicted metadata measures')
  } else {
    message('The following measures were well-predicted by the microbiome using a ', toupper(ml_method), ' model:')
    message(paste(summary_correlations$measure[summary_correlations$Significant], collapse = ', '))
  }
  
  return(list(
    results = summary_correlations,
    feat_importance = summary_feat_importance,
    cv_cor = cv_metrics,
    fold_preds = fold_preds_all
  ))
}
```

Train the model with microbiome features + confounders and only confounders:

```{r}

PRED_FIXED <- c("breed_size_numeric", "Sex_bool", "pool", "total_reads_before_filtering",
                "duplication_rate", "weight_from_vet_or_survey_lbs")

# microbiome features + confounders

tmp_sample_map <- sample_map %>% 
  filter(!is.na(weight_from_vet_or_survey_lbs))  %>%
  filter(!is.na(breed_size_numeric))  %>%
  filter(hs_general_health %in% c(1,2)) %>%
  mutate(Sex_bool = case_when(
    grepl("Female", Sex_class) ~ "female",
    grepl("Male", Sex_class) ~ "male",
    TRUE ~ Sex_class # This is a catch-all in case there are unexpected values
  ))

tmp_ra_kraken_sp <- ra_kraken_sp %>% 
  inner_join(tmp_sample_map %>% select(PRED_FIXED, sample_id), by="sample_id") 

rf_res_mic <- get_predictability_regression_for_age_pred(tmp_ra_kraken_sp, 
                                                         tmp_sample_map, 
                                                         "sample_id", 
                                                         "age_at_stool_collection", 
                                                         ml_method = 'rf',
                                                         fdr_threshold = 1,
                                                         corr_threshold = 0,
                                                         corr_method = "pearson",
                                                         n_ml_repeats = 1, 
                                                         n_ml_folds = 10, 
                                                         feat_importance = 'altmann',
                                                         rand_seed = 84,
                                                         quiet = FALSE)

# only confounders

confounders <- tmp_sample_map %>% select(sample_id, PRED_FIXED) %>%
  mutate(breed_size_numeric = as.numeric(breed_size_numeric))

rf_res_con <- get_predictability_regression_for_age_pred(confounders, 
                                          tmp_sample_map, 
                                          "sample_id", 
                                          "age_at_stool_collection", 
                                          ml_method = 'rf',
                                          fdr_threshold = 1,
                                          corr_threshold = 0,
                                          corr_method = "pearson",
                                          n_ml_repeats = 1, 
                                          n_ml_folds = 10, 
                                          feat_importance = 'none',
                                          rand_seed = 84,
                                          quiet = FALSE)


final_df_mic <- do.call(rbind, lapply(seq_along(rf_res_mic$fold_preds), function(i) {
  df <- rf_res_mic$fold_preds[[i]]
  df$index <- i
  df$data <- "microbiome + confounders"
  return(df)
}))

final_df_con <- do.call(rbind, lapply(seq_along(rf_res_con$fold_preds), function(i) {
  df <- rf_res_con$fold_preds[[i]]
  df$index <- i
  df$data <- "confounders"
  return(df)
}))

merged_df <- rbind(final_df_mic, final_df_con)

```

Plot the results of the models:

```{r fig.height=4, fig.width=4, message=FALSE, warning=FALSE}

library(ggpmisc)

# Calculate correlations per facet
correlations <- merged_df %>%
  filter(data == "microbiome + confounders") %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds)) %>%
  mutate(data="microbiome + confounders")
correlations2 <- merged_df %>%
  filter(data == "confounders") %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds)) %>%
  mutate(data="confounders")

merged_correlations <- rbind(correlations, correlations2)

p <- ggplot(merged_correlations, aes(x = data, y = correlation)) +
  geom_boxplot(aes(fill=data), alpha=0.5, show.legend = FALSE) +
  geom_point(size=2.5, alpha=0.5) + #, aes(color=as.factor(index))
  labs(x = "Training data", y = "Perason correlation") + # , color = "Fold"
  # scale_color_manual(values = c("#012a4a", "#013a63", "#01497c", "#014f86", "#2a6f97", "#2c7da0", "#468faf", "#61a5c2", "#89c2d9", "#a9d6e5")) +
  scale_fill_manual(values = c("#b5e48c", "#89c2d9")) +
  theme_classic() +
  theme(
    panel.grid.major = element_line(color = "lightgray", size = 0.25),  # Major grid lines
    panel.grid.minor = element_line(color = "lightgray", size = 0.25),  # Minor grid lines
    panel.grid.major.x = element_blank(),  # Remove vertical grid lines
    panel.grid.minor.x = element_blank(),  # Remove minor vertical grid lines
  ) +
  scale_x_discrete(labels = c("microbiome + confounders" = "Microbiome + confounders", "confounders" = "Confounders"))  # Manually set x-axis labels

print(p) # Figure 3G

```

Plot the correlation between the real and predicted age per fold (for the model that used microbiome features + confounders):

```{r fig.height=4, fig.width=5, message=FALSE, warning=FALSE}
# Load necessary packages
library(ggplot2)
library(dplyr)
library(ggpubr)

data <- merged_df %>% filter(data == "microbiome + confounders")

# Calculate Spearman correlations for each index
correlations <- data %>%
  group_by(index) %>%
  summarize(correlation = cor(true_label, preds, method = "pearson")) %>%
  mutate(label = paste(index))#, ": Pearson r =", round(correlation, 3)))

# Convert correlations to a named vector for legend labels
legend_labels <- setNames(correlations$label, as.character(correlations$index))

# Create the scatter plot with regression lines
p <- ggplot(data, aes(x = true_label, y = preds)) +
  geom_point(size=1.25, alpha=0.3) +
  geom_smooth(aes(color = as.factor(index)), method = "lm", se = FALSE, size=0.5, alpha=0.3) +
  labs(color = "Fold") +
  xlab("Real age") +
  ylab("Predicted age") +
  scale_color_manual(values = c("#012a4a", "#013a63", "#01497c", "#014f86", "#2a6f97", "#2c7da0", "#468faf", "#61a5c2", "#89c2d9", "#a9d6e5"), 
                     labels = legend_labels) +
  theme_classic()

# Display the plot
print(p) # Figure 3F

```


Spearman correlation between MAASLIN2 results and feature importance

```{r}

feat_imp <- rf_res_mic$feat_importance


p_a <- feat_imp %>%
  filter(fdr < 0.05) %>%
  arrange(fdr) %>%
  # head(30) %>%
  ggplot(aes(x = reorder(feature, -mean_feat_importance), y = mean_feat_importance)) +
  geom_bar(stat = "identity", fill = "#2a6f97", alpha=0.5, color="black", size=0.2) +
  theme_classic(base_size = 14) +
  theme(axis.text.y = element_text(angle = 0, hjust = 1)) +  # Adjust angle as needed
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +   # Adjust angle as needed)
  labs(x = "Feature", y = "Mean Feature Importance")

p_a # Supplementary Figure S8A

df <- feat_imp %>% select(feature, mean_feat_importance) %>%
  inner_join(maaslin_res %>% 
               filter(metadata == "age_at_stool_collection") %>%
               select(feature, coef) %>% rename(maaslin_coef = coef), by="feature") %>%
  mutate(abs_maaslin_coef = abs(maaslin_coef)) 
  
  # summarise(corr = cor(mean_feat_importance, abs_maaslin_coef, method = "spearman")) 
  
p_b <- ggplot(df, aes(x = mean_feat_importance, y = abs_maaslin_coef)) +
  geom_point(color = 'black', shape = 21, size = 3, alpha = 0.7, fill = "gray80") +  # was: color_by_var_
  geom_smooth(method = 'lm', se=FALSE, color="darkred") +
  theme_classic(base_size = 14) +
  xlab("Mean Altmann feature importance (RF)") +
  ylab("Absolute Maaslin2 Coefficient")

p_b # Supplementary Figure S8B

```

```{r}
df %>%
  summarise(corr = cor(mean_feat_importance, abs_maaslin_coef, method = "spearman"))

res <- with(df, cor.test(mean_feat_importance, abs_maaslin_coef,
                         method = "spearman", exact = FALSE))
res$estimate   # Spearman rho
res$p.value    # p-value

```

Test the model on other dataset

```{r}

# Read Coehlo et al. 2018 data

CO_FOLDER <- "../data/coehlo_et_al_2018"

coe_metadata <- read_csv(file.path(CO_FOLDER,"table_s1.csv"), show_col_types = F)
reads <- read_csv(file.path(CO_FOLDER,"table_s2.csv"), show_col_types = F)

coe_metadata <- coe_metadata %>%
  filter(Diet == 'Baseline') %>%
  left_join(reads, by = "sample_id") %>%
  select(-sample_id, -Diet, -`Final fecal date`, -Note) %>%
  rename(sample_id = ENA_ACCESSION) %>%
  relocate(sample_id)

rm(reads)

kraken_sp_coe <- read_tsv2(
  file.path(
    CO_FOLDER, 
    "feature_tables",
    "kraken_species_level_taxonomy_RA.tsv")) %>%
  reorganize_kraken_table()

kraken_ge_coe <- read_tsv2(
  file.path(
    CO_FOLDER, 
    "feature_tables",
    "kraken_genus_level_taxonomy_RA.tsv")) %>%
  reorganize_kraken_table()

kraken_ph_coe <- read_tsv2(
  file.path(
    CO_FOLDER, 
    "feature_tables",
    "kraken_phylum_level_taxonomy_RA.tsv")) %>%
  reorganize_kraken_table()

message('Converting Kraken tables to relative abundance')
ra_kraken_ph_coe <- convert_to_relab(kraken_ph_coe)
ra_kraken_ge_coe <- convert_to_relab(kraken_ge_coe)
ra_kraken_sp_coe <- convert_to_relab(kraken_sp_coe)

prevalence_cutoff = 0.2
avg_abundance_cutoff = 0.0001

message('Removing rare phyla from Kraken tables')
ra_kraken_ph_coe <- remove_rare_features(
  ra_kraken_ph_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Removing rare genera from Kraken tables')
ra_kraken_ge_coe <- remove_rare_features(
  ra_kraken_ge_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Removing rare species from Kraken tables')
ra_kraken_sp_coe <- remove_rare_features(
  ra_kraken_sp_coe,
  prevalence_cutoff=prevalence_cutoff,
  avg_abundance_cutoff=avg_abundance_cutoff)

message('Re-normalizing to relative abundances')
ra_kraken_ph_coe <- convert_to_relab(ra_kraken_ph_coe)
ra_kraken_ge_coe <- convert_to_relab(ra_kraken_ge_coe)
ra_kraken_sp_coe <- convert_to_relab(ra_kraken_sp_coe)

qc_stats_coe <- read_tsv2(file.path(CO_FOLDER, "feature_tables/qc_stats.tsv"))

```


```{r}
get_predictability_regression_with_test <- function(feat_table, 
                                          metadata, 
                                          sample_id_column, 
                                          vars_to_analyze, 
                                          ml_method = 'rf',
                                          fdr_threshold = 0.1,
                                          corr_threshold = 0.3,
                                          corr_method = "pearson",
                                          n_ml_repeats = 10,
                                          n_ml_folds = 5,
                                          feat_importance = 'none',
                                          rand_seed = 1111,
                                          quiet = FALSE,
                                          # --- external test data ---
                                          test_feat_table = NULL,
                                          test_metadata   = NULL) {
  # Required libraries
  require(rsample)
  require(dplyr)
  require(glmnet)
  require(ranger)
  require(metap)
  require(tibble)

  # --- validations ---
  if (is.null(vars_to_analyze)) {
    stop("Invalid 'vars_to_analyze' argument. Cannot be NULL")
  }
  if (!all(vars_to_analyze %in% names(metadata))) {
    stop("Invalid 'vars_to_analyze': some variables are missing from training metadata")
  }
  if (!ml_method %in% c('rf')) {
    stop("Invalid 'ml_method'. Only 'rf' supported.")
  }
  if (!corr_method %in% c('pearson', 'spearman')) {
    stop("Invalid 'corr_method'. Only 'pearson' and 'spearman' supported.")
  }

  has_external <- !is.null(test_feat_table) && !is.null(test_metadata)

  # Fix names of features
  names(feat_table) <- make.names(names(feat_table))
  if (has_external) {
    names(test_feat_table) <- make.names(names(test_feat_table))
  }

  # Folds for CV
  n_samples <- nrow(feat_table)
  set.seed(rand_seed)
  folds <- vfold_cv(
    data.frame(sample_num = 1:n_samples),
    v = n_ml_folds,
    repeats = n_ml_repeats
  )

  # Collectors
  cv_correlations          <- data.frame()
  cv_feat_imp              <- data.frame()
  all_fold_preds_internal  <- list()
  all_fold_preds_external  <- list()
  cv_correlations_external <- data.frame()

  # helper for p-value aggregation
  sumlog2 <- function(ps) {
    ps <- ps[!is.na(ps)]
    if (!length(ps)) return(NA_real_)
    sumlog(ps)$p
  }

  # regression metrics helper
  calc_regression_metrics <- function(preds, true_vals, corr_method = "pearson") {
    keep <- stats::complete.cases(preds, true_vals)
    preds <- preds[keep]
    true_vals <- true_vals[keep]

    if (length(preds) == 0) {
      return(list(
        cor = NA_real_,
        cor_p = NA_real_,
        rmse = NA_real_,
        mae = NA_real_,
        r2 = NA_real_
      ))
    }

    cor_val <- suppressWarnings(cor(preds, true_vals, method = corr_method))

    cor_p_val <- tryCatch(
      suppressWarnings(
        cor.test(
          preds, true_vals,
          alternative = "greater",
          method = corr_method,
          exact = FALSE
        )$p.value
      ),
      error = function(e) NA_real_
    )

    rmse_val <- sqrt(mean((preds - true_vals)^2, na.rm = TRUE))
    mae_val  <- mean(abs(preds - true_vals), na.rm = TRUE)

    ss_res <- sum((true_vals - preds)^2, na.rm = TRUE)
    ss_tot <- sum((true_vals - mean(true_vals, na.rm = TRUE))^2, na.rm = TRUE)

    r2_val <- if (is.na(ss_tot) || ss_tot == 0) NA_real_ else 1 - (ss_res / ss_tot)

    list(
      cor = cor_val,
      cor_p = cor_p_val,
      rmse = rmse_val,
      mae = mae_val,
      r2 = r2_val
    )
  }

  # Iterate over variables
  for (measure in vars_to_analyze) {
    if (!quiet) {
      message("Training model for variable: ", measure)
    }

    # Join label into training features
    tmp <- feat_table %>%
      left_join(
        metadata[, c(sample_id_column, measure)] %>% rename(Label = 2),
        by = sample_id_column
      )

    # External test set prepared once per measure
    if (has_external) {
      tmp_ext <- test_feat_table %>%
        left_join(
          test_metadata[, c(sample_id_column, measure)] %>% rename(Label = 2),
          by = sample_id_column
        )
    }

    pred_per_fold_internal <- list()
    pred_per_fold_external <- list()

    # CV loop
    for (i in seq_len(nrow(folds))) {
      fold_id <- paste(folds$splits[[i]]$id, collapse = " ")

      # Train/holdout split
      train_idx <- folds$splits[[i]]$in_id
      train_data_full <- tmp[train_idx, , drop = FALSE] %>% filter(!is.na(Label))
      test_data_full  <- tmp[-train_idx, , drop = FALSE] %>% filter(!is.na(Label))

      # Keep IDs
      test_ids_internal <- test_data_full[[sample_id_column]]

      # Drop ID column for modeling
      train_data <- train_data_full %>% select(-any_of(sample_id_column))
      test_data  <- test_data_full  %>% select(-any_of(sample_id_column))

      # Features
      train_feats <- setdiff(names(train_data), "Label")

      # Train RF
      rf_model <- ranger(
        Label ~ ., data = train_data,
        importance = ifelse(feat_importance == 'altmann', 'permutation', 'none'),
        seed = rand_seed
      )

      if (feat_importance != 'none') {
        feat_imp <- importance_pvalues(
          rf_model, method = "altmann", formula = Label ~ .,
          data = train_data, seed = rand_seed
        )
        cv_feat_imp <- bind_rows(
          cv_feat_imp,
          as.data.frame(feat_imp) %>%
            tibble::rownames_to_column('feature') %>%
            mutate(measure = measure, fold_id = fold_id)
        )
      }

      # Predict on internal holdout
      preds_int <- predict(rf_model, test_data %>% select(all_of(train_feats)))$predictions
      df_int <- data.frame(
        measure    = measure,
        fold_id    = fold_id,
        sample_id  = test_ids_internal,
        preds      = preds_int,
        true_label = test_data$Label,
        stringsAsFactors = FALSE
      )
      pred_per_fold_internal[[i]] <- df_int

      # Internal metrics
      int_metrics <- calc_regression_metrics(
        preds = df_int$preds,
        true_vals = df_int$true_label,
        corr_method = corr_method
      )

      cv_correlations <- bind_rows(
        cv_correlations,
        data.frame(
          measure    = measure,
          fold_id    = fold_id,
          test_cor   = int_metrics$cor,
          test_cor_p = int_metrics$cor_p,
          rmse       = int_metrics$rmse,
          mae        = int_metrics$mae,
          r2         = int_metrics$r2
        )
      )

      # Predict on external test set
      if (has_external) {
        ext_keep <- intersect(names(tmp_ext), c(sample_id_column, train_feats, "Label"))
        ext_data_full <- tmp_ext[, ext_keep, drop = FALSE] %>% filter(!is.na(Label))

        missing_feats <- setdiff(train_feats, names(ext_data_full))
        if (length(missing_feats) > 0) {
          for (mf in missing_feats) {
            ext_data_full[[mf]] <- 0
          }
        }

        ext_feats <- ext_data_full %>% select(all_of(train_feats))
        ext_ids   <- ext_data_full[[sample_id_column]]
        preds_ext <- predict(rf_model, ext_feats)$predictions

        df_ext <- data.frame(
          measure    = measure,
          fold_id    = fold_id,
          sample_id  = ext_ids,
          preds      = preds_ext,
          true_label = ext_data_full$Label,
          stringsAsFactors = FALSE
        )
        pred_per_fold_external[[i]] <- df_ext

        ext_metrics <- calc_regression_metrics(
          preds = df_ext$preds,
          true_vals = df_ext$true_label,
          corr_method = corr_method
        )

        cv_correlations_external <- bind_rows(
          cv_correlations_external,
          data.frame(
            measure    = measure,
            fold_id    = fold_id,
            test_cor   = ext_metrics$cor,
            test_cor_p = ext_metrics$cor_p,
            rmse       = ext_metrics$rmse,
            mae        = ext_metrics$mae,
            r2         = ext_metrics$r2
          )
        )
      }
    }

    # store per-measure predictions
    all_fold_preds_internal[[measure]] <- dplyr::bind_rows(pred_per_fold_internal)
    if (has_external) {
      all_fold_preds_external[[measure]] <- dplyr::bind_rows(pred_per_fold_external)
    }
  }

  # Summaries (internal CV)
  summary_correlations <- cv_correlations %>%
    group_by(measure) %>%
    summarise(
      n_folds   = n(),
      mean_cor  = mean(test_cor, na.rm = TRUE),
      sd_cor    = sd(test_cor, na.rm = TRUE),
      mean_rmse = mean(rmse, na.rm = TRUE),
      sd_rmse   = sd(rmse, na.rm = TRUE),
      mean_mae  = mean(mae, na.rm = TRUE),
      sd_mae    = sd(mae, na.rm = TRUE),
      mean_r2   = mean(r2, na.rm = TRUE),
      sd_r2     = sd(r2, na.rm = TRUE),
      p         = sumlog2(test_cor_p),
      .groups   = "drop"
    ) %>%
    mutate(
      fdr = p.adjust(p),
      Significant = fdr < fdr_threshold & mean_cor > corr_threshold
    )

  # Summaries (external test)
  summary_correlations_external <- NULL
  if (has_external) {
    summary_correlations_external <- cv_correlations_external %>%
      group_by(measure) %>%
      summarise(
        n_folds   = n(),
        mean_cor  = mean(test_cor, na.rm = TRUE),
        sd_cor    = sd(test_cor, na.rm = TRUE),
        mean_rmse = mean(rmse, na.rm = TRUE),
        sd_rmse   = sd(rmse, na.rm = TRUE),
        mean_mae  = mean(mae, na.rm = TRUE),
        sd_mae    = sd(mae, na.rm = TRUE),
        mean_r2   = mean(r2, na.rm = TRUE),
        sd_r2     = sd(r2, na.rm = TRUE),
        p         = sumlog2(test_cor_p),
        .groups   = "drop"
      ) %>%
      mutate(
        fdr = p.adjust(p),
        Significant = fdr < fdr_threshold & mean_cor > corr_threshold
      )
  }

  # Feature importance summary
  summary_feat_importance <- data.frame()
  if (feat_importance != 'none' && nrow(cv_feat_imp) > 0) {
    summary_feat_importance <- cv_feat_imp %>%
      group_by(measure, feature) %>%
      summarise(
        n_folds = n(),
        mean_feat_importance = mean(importance, na.rm = TRUE),
        sd_feat_importance   = sd(importance, na.rm = TRUE),
        p = sumlog2(pvalue),
        .groups = "drop"
      ) %>%
      mutate(fdr = p.adjust(p))
  }

  if (!any(summary_correlations$Significant)) {
    message("No significantly well-predicted metadata measures (internal CV).")
  } else {
    message(
      "Well-predicted (internal CV, ", toupper(ml_method), "): ",
      paste(summary_correlations$measure[summary_correlations$Significant], collapse = ", ")
    )
  }

  if (has_external && any(summary_correlations_external$Significant)) {
    message(
      "Well-predicted on external test: ",
      paste(summary_correlations_external$measure[summary_correlations_external$Significant], collapse = ", ")
    )
  }

  return(list(
    results              = summary_correlations,
    results_external     = summary_correlations_external,
    feat_importance      = summary_feat_importance,
    cv_cor               = cv_correlations,
    cv_cor_external      = cv_correlations_external,
    fold_preds           = all_fold_preds_internal,
    fold_preds_external  = if (has_external) all_fold_preds_external else NULL
  ))
}

```


```{r}


CO_PRED_FIXED <- c("Sex_bool", "total_reads_before_filtering", "duplication_rate")

tmp_sample_map <- sample_map %>% 
  filter(!is.na(weight_from_vet_or_survey_lbs))  %>%
  filter(!is.na(breed_size_numeric))  %>%
  filter(hs_general_health %in% c(1,2)) %>%
  mutate(Sex_bool = case_when(
    grepl("Female", Sex_class) ~ "F",
    grepl("Male", Sex_class) ~ "M",
    TRUE ~ Sex_class # This is a catch-all in case there are unexpected values
  ))


tmp_ra_kraken_sp_coe <-  ra_kraken_sp_coe %>% 
  inner_join(coe_metadata %>% select(sample_id, Sex), by="sample_id") %>%
  rename(Sex_bool = Sex) %>%
  inner_join(qc_stats_coe %>% select(sample_id, total_reads_before_filtering, duplication_rate), by="sample_id") 

tmp_coe_metadata <- coe_metadata %>%
  filter(Phenotype != 'Overweight to obese') 

rf_res_co <- get_predictability_regression_with_test(ra_kraken_sp, 
                                          tmp_sample_map, 
                                          "sample_id", 
                                          "age_at_stool_collection", 
                                          ml_method = 'rf',
                                          fdr_threshold = 1,
                                          corr_threshold = 0,
                                          corr_method = "pearson",
                                          n_ml_repeats = 1, # was 10
                                          n_ml_folds = 10, # was 5
                                          feat_importance = 'none',
                                          rand_seed = 84,
                                          quiet = FALSE,
                                          # --- external test data ---
                                          test_feat_table = ra_kraken_sp_coe,
                                          test_metadata   = tmp_coe_metadata %>%
                                            rename(age_at_stool_collection = Age))

```

```{r, fig.height=4, fig.width=4, message=FALSE, warning=FALSE}}
# Load necessary packages
library(ggplot2)
library(dplyr)
library(ggpubr)

data <- final_df_mic <- do.call(rbind, lapply(seq_along(rf_res_co$fold_preds_external), function(i) {
  df <- rf_res_co$fold_preds_external[[i]]
  return(df)
})) %>%
  mutate(fold_index = as.numeric(gsub("Fold", "", fold_id)))


p <- ggplot(data, aes(x = true_label, y = preds)) +
  geom_point(size=1.25, alpha=0.3) +
  geom_smooth(aes(group = fold_index, color = factor(fold_index)),
              method = "lm", se = FALSE, size = 0.5, alpha = 0.3) +
  xlab("Real age") +
  ylab("Predicted age") +
  labs(color = "Fold") +
  theme_classic(base_size = 14) +
    scale_color_manual(values = c(
  "#2d004b",
  "#4b007d",
  "#5a189a",
  "#6a23b5",
  "#7b2cbf",
  "#9254d5",
  "#a86fe9",
  "#c77dff",
  "#d9a8ff",
  "#e0aaff"
)) 

p # Figure 3H
```


 
