---
title: "High-level associations between omics and measurement types"
output:
  html_document:
    df_print: paged
    code_folding: hide
---

Here we aimed to assess the global associations between the dog gut microbiome and multiple technical and biological measurements available as part of the DAP. These measurement types included technical factors (such as sequencing batch and sequencing quality metrics), dog demographics, and lifestyle information as reported by dog owners. We additionally included complete blood counts (CBC) and genetic profiles (see Methods). A full list of measurement types and the features included within them is given in Supplementary Table S2. 

Preparations:

```{r message=FALSE, warning=FALSE}
# Preparations
library(tidyverse)
library(readr)
library(ggplot2)
library(tibble)
library(tidyr)
library(dplyr)
library(ape)
library(cowplot)
library(vegan)
library(ecodist)
library(stringr)
library(foreach)
library(variancePartition)
library(PMA)
library(glmnet)

# Setup local cluster for parallel computing
library(doSNOW)
cl <- makeCluster(6)
registerDoSNOW(cl)

# Resolve possible conflicts
library(conflicted)
filter <- dplyr::filter
select <- dplyr::select

options(scipen = 999)
set.seed(217)
source('utils.R')

# Paths
metadata_dir <- '../data/metadata'
terra_release <- 'terra_release_2023_12022024'
dog_and_owner_csv <- file.path(metadata_dir, terra_release, "DAP_2023_HLES_dog_owner_v1.0.csv")
metadata_codebook <- file.path(metadata_dir, terra_release, "DAP_2023_CODEBOOK_v1.0.csv")
dap_breeds <- file.path(metadata_dir, terra_release, "DogAgingProject_GeneticData_CuratedRelease_2023.ancestry.tsv")
dap_genetic_distance_matrix <- "../data/metadata/prec_grm.rds"
dap_lab_tests <- file.path(metadata_dir, terra_release, "DAP_2023_SamplesResults_CBC_v1.0.csv")
dir.create("omic_associations_backup")
dir.create("Figure2")
dir.create("SuppFigure2")

# Mark whether all tests should be re-run, or whether previously-saved backup file can be used
RECALCULATE_TESTS <- FALSE
```

## Load DAP profiles

```{r DAP_DATA_LOADING, warning=FALSE, message=FALSE, cache = TRUE}
source('../raw_shotgun_processing/load_batches_1-6.R') 
```

```{r, cache=TRUE}
# Read metadata codebook
codebook <- read_csv(metadata_codebook, show_col_types = FALSE, na = c('NA'))

# Read main HLES table
dog_and_owner <- read_csv(dog_and_owner_csv, show_col_types = FALSE, guess_max = 50000) %>%
  rename(dap_dog_id = dog_id) %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  mutate(dap_dog_id = as.character(dap_dog_id))
```

***

### Utility functions

#### Plotting utils

```{r}
get_signif_marks <- function(ps) {
  sapply(ps,
         function(p) {
           if (p<=0.005) return("***")
           if (p<=0.01) return("**")
           if (p<=0.05) return("*")
           if (p<=0.1) return(".")
           return("")
         })
}
```

```{r}
# For Mantel/Procrustes
plot_pairwise_shared_var <- function(df, title, effect_size_title, var_exp_cut = 5) {
  df <- df %>% 
    mutate(lbl = ifelse(
      sign == "", "", 
      paste0(sign,'\n',round(var_explained,1),'%')
      ))
  
  # Push microbiome to top of the matrix
  microb_dist <- grep("Microbiome", unique(df$dist_x_pretty), value = T)
  dm_order <- rev(unique(c(
    microb_dist[1], 
    sort(unique(df$dist_x_pretty[! df$dist_x_pretty %in% microb_dist])),
    microb_dist[2]
    )))
  df$dist_x_pretty <- factor(df$dist_x_pretty, levels = dm_order)
  df$dist_y_pretty <- factor(df$dist_y_pretty, levels = dm_order)
  
  # So that extremely high associations are not overshadowing all else
  df$var_explained <- pmin(var_exp_cut, df$var_explained)
  
  p <- ggplot(df %>%
           filter(as.numeric(dist_x_pretty) <= as.numeric(dist_y_pretty)), 
         aes(x = dist_x_pretty, y= dist_y_pretty, fill = var_explained)) +
    geom_tile(color = "black", linewidth = 0.5, alpha = 0.6) +
    geom_text(aes(label = lbl), color = "black", size = 2.6, lineheight = .7) +
    geom_rect(xmin = 0.5, xmax = length(dm_order)-0.5, 
              ymin = length(dm_order)-1.5, ymax = length(dm_order)-0.5, 
              size = 0.8, color = 'black', alpha = 0) +
    geom_rect(ymin = 0.5, ymax = length(dm_order)-0.5, 
              xmin = 0.5, xmax = 1.5, 
              size = 0.8, color = 'black', alpha = 0) +
    theme_minimal(base_size = 13) +
    labs(x = NULL, y = NULL) +
    ggtitle(title) +
    scale_x_discrete(position = "top") +
    scale_fill_gradientn(colors = c('grey90','#afcb90',
                                    '#8FB7B2','#707E9E',
                                    '#4A618B','#2c497f',
                                    '#3d2b56'),
                         breaks = 0:var_exp_cut,
                         labels = c(0:(var_exp_cut-1), paste0('>',var_exp_cut)),
                         name = effect_size_title,  
                         guide = guide_colorbar(frame.colour = "black", 
                                                ticks.colour = "black")) +
    theme(plot.title = element_text(hjust = 0.5, size = 14)) +
    theme(panel.grid.major = element_blank()) +
    theme(axis.text.x.top = element_text(angle = 45, hjust = 0)) +
    theme(legend.position = c(.84, .23))
  
  return(p)
}
```

#### Distance matrices

```{r}
get_dist_mat <- function(mat, dist_method = "euclidean") {
  # For purely binary data
  if (dist_method == 'binary') {
    mat_dist <- dist(mat, method = 'binary') %>% as.matrix()
  
  # Euclidean distances
  } else if (dist_method == 'euclidean') {
    # Normalize variables
    mat_centered <- scale(mat, center = TRUE, scale = TRUE)
    # Calculate distances
    mat_dist <- dist(mat_centered, method = dist_method) %>% as.matrix()
  } else {
    stop('Invalid distance metric')
  }
  return(mat_dist)
}

# Wrapper to above function that takes a data frame with sample ID's column as input
get_dist_mat_from_df <- function(df, dist_method = "euclidean", sample_id_column = 'dap_dog_id') {
  return(
    df %>%
      column_to_rownames(sample_id_column) %>%
      as.matrix() %>%
      get_dist_mat(dist_method = dist_method) %>%
      as_tibble(rownames = sample_id_column, .name_repair = "minimal")
  )
}
```

```{r}
get_categorical_dist <- function(feat_table, 
                                 cat_column, 
                                 sample_id_column = 'dap_dog_id') {
  n_dogs <- nrow(feat_table)
  cat_mat <- matrix(0, nrow = n_dogs, ncol = n_dogs)
  for (i in 1:(n_dogs-1)) {
    for (j in (i+1):n_dogs) {
      if(feat_table[[cat_column]][i] != feat_table[[cat_column]][j]) {
        cat_mat[i,j] <- 1
        cat_mat[j,i] <- 1
      }
    }
  }
  rownames(cat_mat) <- feat_table[[sample_id_column]]
  colnames(cat_mat) <- feat_table[[sample_id_column]]
  cat_mat <- as.data.frame(cat_mat) %>% rownames_to_column(sample_id_column)
  return(cat_mat)
}
```

***

### Data processing

***

#### Microbiome - taxonomy

```{r}
bc_dap_dog_id <- map_sample_id_to_dap_dog_id(bc, sample_map = sample_map)
message('Done preparing bray-curtis distances')

wuf_dap_dog_id <- map_sample_id_to_dap_dog_id(wuf, sample_map = sample_map)
message('Done preparing weighted unifrac distances')

shannon <- map_sample_id_to_dap_dog_id(shannon, sample_map = sample_map, rename_cols = FALSE)
ra_kraken_sp <- map_sample_id_to_dap_dog_id(ra_kraken_sp, sample_map, rename_cols = F)
ra_kraken_ge <- map_sample_id_to_dap_dog_id(ra_kraken_ge, sample_map, rename_cols = F)
ra_kraken_ph <- map_sample_id_to_dap_dog_id(ra_kraken_ph, sample_map, rename_cols = F)
```

***

#### Microbiome - function

```{r}
# Compute distance matrices based on MetaLAFFA functional profiles
kegg_pwys <- map_sample_id_to_dap_dog_id(
  kegg_pwys, 
  sample_map = sample_map, 
  rename_cols = FALSE)

# Calculate distance matrices
kegg_pwy_dist <- get_dist_mat_from_df(kegg_pwys)

message('Done preparing functional data')
```

***

#### Genetic data

```{r}
genetic_distance_matrix <- read_rds(dap_genetic_distance_matrix) %>%
  get_genetic_distance_matrix() %>%
  as.data.frame() 

breed_data <- read_delim(
  dap_genetic_matrix,
  delim = "\t", 
  escape_double = FALSE, 
  trim_ws = TRUE,
  show_col_types = FALSE
  ) %>%
  mutate(dog = as.character(dog)) %>%
  mutate(pop = make.names(pop)) %>%
  select(-code) %>%
  # Remove rare breeds
  group_by(pop) %>%
  filter(n() >= 50) %>%
  ungroup() %>%
  # Convert to wide format
  pivot_wider(id_cols = dog, names_from = pop, values_from = pct, values_fill = 0) %>%
  rename(dap_dog_id = dog)

message('Done preparing genetic data')
```

***

#### Lab test data

CBC only.

```{r}
# Read lab test codes
lab_test_codes <- codebook %>%
  filter(DataFile == "SamplesResults_CBC") %>%
  filter(str_detect(Variable, "krt_cbc_") ) %>%
  filter(!str_detect(Variable, "morph") ) %>%
  filter(!str_detect(Variable, "comments") ) %>%
  filter(!str_detect(Variable, "hemoparasites") ) %>%
  filter(!str_detect(Variable, "plt_quant") ) %>%
  pull(Variable)

# Read and filter lab test results
lab_test <- read_csv(dap_lab_tests, show_col_types = FALSE, na = c('NA')) %>%
  # Filter dogs 
  rename(dap_dog_id = dog_id ) %>%
  mutate(dap_dog_id = as.character(dap_dog_id)) %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  # Precision 1 only
  filter(Sample_Year == 'precision_1') %>%
  # Take only numeric lab tests
  select(all_of(c("dap_dog_id", lab_test_codes))) %>%
  # Remove columns that have only zeros or NA
  select_if(~ !all(is.na(.) | . == 0)) 

# Remove columns with more than 20% NA's
lab_test <- lab_test[, which(colMeans(is.na(lab_test)) < 0.2)]

# Remove rows with missing data
lab_test <- lab_test %>% drop_na()

# Calculate distance matrices
lab_test_dist <- get_dist_mat_from_df(lab_test)

message('Done preparing lab test data')
```

***

#### Dog overview - age, weight, sex [`dd`]

```{r}
# Age-only distance matrix
age_data <- sample_map %>%
  select(dap_dog_id, age_at_stool_collection)
age_dist <- age_data %>%
  get_dist_mat_from_df()

# Weight-only distance matrix
weight_data <- sample_map %>%
  select(dap_dog_id, weight_from_vet_or_survey_lbs)
weight_dist <- weight_data %>%
  drop_na() %>%
  get_dist_mat_from_df()

# Sex-only distance matrix
sex_data <- sample_map %>%
  select(dap_dog_id, Sex_bool, Fixed_bool)
sex_dist <- get_categorical_dist(dog_overview, "Sex_Class_at_HLES")
```


***

#### Environment & lifestyle [`de`, `pa`]

`de` = Dog Environment  
`pa` = Physical Activity  

```{r}
env_codes <- c("de_recent_toxins_or_hazards_ingested_frequency", # Toxin exposure
               "de_other_present_animals_dogs",                  # Household
               "de_other_present_animals_cats",
               "de_other_present_animals_wildlife",
               "de_routine_hours_per_day_with_other_animals",
               "de_routine_hours_per_day_with_children",
               "de_routine_hours_per_day_with_adults",
               "pa_cold_weather_months_per_year",                # Weather
               "pa_hot_weather_months_per_year",                 
               "de_property_area_accessible",                    # Property
               "de_air_cleaner_present",                         # Air 
               "de_air_freshener_present",
               "de_hepa_present",
               "de_home_area_type")

pa_codes <- c("pa_avg_daily_active_hours",                       # PA
              "pa_activity_level",                               
              "pa_avg_activity_intensity",                       
              "pa_on_leash_walk_frequency",
              "pa_physical_games_frequency",
              "de_stairs_in_home",
              "de_routine_hours_per_day_in_yard",
              "de_daytime_sleep_avg_hours",                      # Sleep
              "de_nighttime_sleep_avg_hours")

environment_data <- dog_and_owner %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  select(dap_dog_id, all_of(env_codes)) %>%
  mutate(de_property_area_accessible = 
           ifelse(is.na(de_property_area_accessible), 0, de_property_area_accessible)) %>%
  mutate(de_other_present_animals_dogs = 
           ifelse(de_other_present_animals_dogs == "TRUE", 1, 0)) %>%
  mutate(de_other_present_animals_cats = 
           ifelse(de_other_present_animals_cats == "TRUE", 1, 0)) %>%
  mutate(de_other_present_animals_wildlife = 
           ifelse(de_other_present_animals_wildlife == "TRUE", 1, 0)) %>%
  mutate(de_air_cleaner_present = 
           ifelse(de_air_cleaner_present == 1 | de_hepa_present == 1, 1, 0)) %>%
  mutate(de_air_freshener_present = 
           ifelse(de_air_freshener_present < 99, de_air_freshener_present, NA)) %>%
  select(-de_hepa_present) %>%
  mutate(de_routine_hours_per_day_with_other_animals = 
           ifelse(de_routine_hours_per_day_with_other_animals == 97, 0.5,  de_routine_hours_per_day_with_other_animals)) %>%
  mutate(de_routine_hours_per_day_with_children = 
           ifelse(de_routine_hours_per_day_with_children == 97, 0.5,  de_routine_hours_per_day_with_children)) %>%
  mutate(de_routine_hours_per_day_with_adults = 
           ifelse(de_routine_hours_per_day_with_adults == 97, 0.5,  de_routine_hours_per_day_with_adults)) 
  
environment_dist <- environment_data %>%
  drop_na() %>%
  get_dist_mat_from_df()

pa_data <- dog_and_owner %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  select(dap_dog_id, all_of(pa_codes)) %>%
  mutate(de_stairs_in_home = 
           ifelse(de_stairs_in_home == "TRUE", 1, 0)) %>%
  mutate(de_routine_hours_per_day_in_yard = 
           ifelse(de_routine_hours_per_day_in_yard == 97, 0.5,  de_routine_hours_per_day_in_yard)) %>%
  mutate(de_daytime_sleep_avg_hours = 
           ifelse(de_daytime_sleep_avg_hours < 99, de_daytime_sleep_avg_hours, NA)) %>%
  mutate(de_nighttime_sleep_avg_hours = 
           ifelse(de_nighttime_sleep_avg_hours < 99, de_nighttime_sleep_avg_hours, NA)) 

pa_dist <- pa_data %>%
  drop_na() %>%
  get_dist_mat_from_df()

message('Done preparing environment and physical activity data')
```

***

#### Diet

```{r}
diet_codes <- c('df_appetite',
                'df_feedings_per_day',
                'df_primary_diet_component',
                'df_treats_frequency',
                "df_treats_homemade_protein",
                "df_treats_homemade_biscuits",
                "df_treats_bones",
                "df_treats_peanut_butter", 
                "df_treats_pumpkin", 
                "df_treats_rawhide", 
                "df_treats_table_carbs", 
                "df_treats_table_meat", 
                "df_treats_vegetables",
                "df_treats_commercial_biscuits",
                'de_drinking_water_is_filtered',
                'de_eats_feces',
                'de_eats_grass_frequency',
                'df_primary_diet_component_grain_free',
                'df_primary_diet_component_organic',
                "de_drinks_outdoor_water",
                # Only supplements given to at least 50 digs are included. NA's equal FALSE.
                'df_daily_supplements_antiox',
                'df_daily_supplements_chondroitin',
                'df_daily_supplements_fiber',
                'df_daily_supplements_glucosamine',
                'df_daily_supplements_omega3',
                'df_daily_supplements_probiotics',
                'df_daily_supplements_vitamins')

# Collect diet data
diet_data <- dog_and_owner %>%
  select(dap_dog_id, all_of(diet_codes)) %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  # Switch numbers so that order makes sense
  mutate(de_eats_grass_frequency = ifelse(de_eats_grass_frequency > 0,
                                          ifelse(de_eats_grass_frequency == 2, 1, 2),
                                          0)) %>%
  mutate(df_daily_supplements_antiox =      ifelse(is.na(df_daily_supplements_antiox)      | df_daily_supplements_antiox==0, 0, 1)) %>%
  mutate(df_daily_supplements_chondroitin = ifelse(is.na(df_daily_supplements_chondroitin) | 
                                                     df_daily_supplements_chondroitin==0, 0, 1)) %>%
  mutate(df_daily_supplements_fiber =       ifelse(is.na(df_daily_supplements_fiber)       | df_daily_supplements_fiber==0, 0, 1)) %>%
  mutate(df_daily_supplements_glucosamine = ifelse(is.na(df_daily_supplements_glucosamine) | 
                                                     df_daily_supplements_glucosamine==0, 0, 1)) %>%
  mutate(df_daily_supplements_omega3 =      ifelse(is.na(df_daily_supplements_omega3)      | df_daily_supplements_omega3==0, 0, 1)) %>%
  mutate(df_daily_supplements_probiotics =  ifelse(is.na(df_daily_supplements_probiotics)  | 
                                                     df_daily_supplements_probiotics==0, 0, 1)) %>%
  mutate(df_daily_supplements_vitamins =    ifelse(is.na(df_daily_supplements_vitamins)    | 
                                                     df_daily_supplements_vitamins==0, 0, 1)) %>%
  mutate(df_treats_frequency = ifelse(df_treats_frequency == 4, 0.5, df_treats_frequency)) %>%
  mutate(de_drinking_water_is_filtered = ifelse(de_drinking_water_is_filtered < 99, de_drinking_water_is_filtered, NA)) %>%
  mutate(de_eats_feces = ifelse(de_eats_feces < 99, de_eats_feces, NA)) %>%
  mutate(df_primary_diet_component_grain_free = ifelse(df_primary_diet_component_grain_free == "TRUE", 1, 0)) %>%
  mutate(df_primary_diet_component_organic = ifelse(df_primary_diet_component_organic == "TRUE", 1, 0)) %>%
  mutate(df_treats_bones = ifelse(df_treats_bones == "TRUE", 1, 0)) %>%
  mutate(df_treats_pumpkin = ifelse(df_treats_pumpkin == "TRUE", 1, 0)) %>%
  mutate(df_treats_peanut_butter = ifelse(df_treats_peanut_butter == "TRUE", 1, 0)) %>%
  mutate(df_treats_rawhide = ifelse(df_treats_rawhide == "TRUE", 1, 0)) %>%
  mutate(df_treats_table_carbs = ifelse(df_treats_table_carbs == "TRUE", 1, 0)) %>%
  mutate(df_treats_table_meat = ifelse(df_treats_table_meat == "TRUE", 1, 0)) %>%
  mutate(df_treats_vegetables = ifelse(df_treats_vegetables == "TRUE", 1, 0)) %>%
  mutate(df_treats_commercial_biscuits = ifelse(df_treats_commercial_biscuits == "TRUE", 1, 0)) %>%
  mutate(df_treats_homemade_protein = ifelse(df_treats_homemade_protein == "TRUE", 1, 0)) %>%
  mutate(df_treats_homemade_biscuits = ifelse(df_treats_homemade_biscuits == "TRUE", 1, 0)) %>%
  mutate(de_drinks_outdoor_water = ifelse(de_drinks_outdoor_water == "TRUE", 1, 0)) %>%
  # Encode the different primary diets
  mutate(value = 1) %>%
  pivot_wider(names_from = "df_primary_diet_component", values_from = "value", values_fill = 0, names_prefix = 'primary_diet_') 

# Drop primary diet categories that are pretty rare
for (feat in grep('^primary_diet_', names(diet_data), value = T)) {
  if (min(table(diet_data[[feat]])) < 30) {
    message('Removed ', feat)
    diet_data[[feat]] <- NULL
  }
}

# Calculate diet distance matrix
diet_dist <- diet_data %>%
  drop_na() %>%
  get_dist_mat_from_df()

message('Done preparing diet data')
```

***

#### Medications (currently only dental care and non-prescription-meds)

```{r}
med_codes <- c("mp_dental_brushing_frequency",
               "mp_dental_treat_frequency",
               "mp_flea_and_tick_treatment_oral",
               "mp_heartworm_preventative_oral_chewable",
               "mp_recent_non_prescription_meds_antihistamine",
               "mp_recent_non_prescription_meds_anti_inflammatory",
               "mp_recent_non_prescription_meds_upset_stomach")

meds_data <- dog_and_owner %>%
  select(dap_dog_id, all_of(med_codes)) %>%
  filter(dap_dog_id %in% dog_overview$dap_dog_id) %>%
  mutate(mp_flea_and_tick_treatment_oral = ifelse(is.na(mp_flea_and_tick_treatment_oral) | mp_flea_and_tick_treatment_oral == "FALSE", 0, 1)) %>%
  mutate(mp_heartworm_preventative_oral_chewable = ifelse(is.na(mp_heartworm_preventative_oral_chewable) | mp_heartworm_preventative_oral_chewable == "FALSE", 0, 1)) %>%
  mutate(mp_recent_non_prescription_meds_antihistamine = ifelse(mp_recent_non_prescription_meds_antihistamine == "TRUE", 1, 0)) %>%
  mutate(mp_recent_non_prescription_meds_anti_inflammatory = ifelse(mp_recent_non_prescription_meds_anti_inflammatory == "TRUE", 1, 0)) %>%
  mutate(mp_recent_non_prescription_meds_upset_stomach = ifelse(mp_recent_non_prescription_meds_upset_stomach == "TRUE", 1, 0)) 

message('Done preparing medication data')
```

***

#### Clinical conditions

In the "dog and owner" table, each health category has an indication out of:
(0) No disorder(s)  | (1) Only congenital disorder(s)  |  (2) Only non-congenital disorder(s)  |  (3) Both congenital and non-congenital disorder(s)

For simplicity, we convert each category into a binary variable (0 = no condition, 1 = otherwise).

```{r} 
# Clinical conditions 
clinical_data <- dog_and_owner %>%
  select(dap_dog_id, starts_with('hs_health_conditions_'), 'hs_general_health') %>%
  select(-hs_health_conditions_other) %>%
  mutate(across(starts_with("hs_health_conditions_"), .fns = ~ case_when(.x == 0 ~ 0, TRUE ~ 1))) %>%
  mutate(is_excellent_health = ifelse(hs_general_health==1, 1, 0)) %>%
  select(-hs_general_health)

clinical_dist <- clinical_data %>%
  get_dist_mat_from_df(dist_method = 'binary')

message('Done preparing clinical data (owner reported)')
```

***

#### Technical variables

Including:  

* Pool   
* Total reads - raw
* Duplication rate
* Q30 bases after quality control

```{r}
sample_qc_data <- qc_stats %>% 
  filter(sample_id %in% sample_map$sample_id) %>%
  select(sample_id, 
         total_reads_before_filtering, 
         duplication_rate,
         q30_bases_rate_after_qc
         ) %>%
  # To reduce effect of outliers
  mutate(total_reads_before_filtering = log10(total_reads_before_filtering))
  
sample_qc_data <- map_sample_id_to_dap_dog_id(
  sample_qc_data, 
  sample_map, 
  rename_cols = F)

# Distance matrix
sample_qc_dist <- sample_qc_data %>%
  drop_na() %>%
  get_dist_mat_from_df()

message('Done preparing techinical confounding data')
```

***

#### Save

Save an R object with the pre-processed data for use by other notebooks.

```{r}
# Organize main demographic variables 
main_demographics <- sample_map %>%
  select(dap_dog_id, 
         age_at_stool_collection, 
         life_stage_at_stool_collection, 
         weight_from_vet_or_survey_lbs,
         Sex_bool,
         Fixed_bool,
         Breed_status,
         Breed,
         breed_size) %>%
  rename(Age = age_at_stool_collection) %>%
  rename(Weight_lbs = weight_from_vet_or_survey_lbs) %>%
  rename(Sex = Sex_bool) %>%
  mutate(Fixed = Fixed_bool) 

save(sample_map,
     ra_kraken_sp,
     ra_kraken_ge,
     ra_kraken_ph,
     kegg_pwys,
     breed_data,
     lab_test,
     main_demographics,
     pa_data,
     environment_data,
     sample_qc_data,
     meds_data,
     diet_data,
     clinical_data,
     file = 'backup_data.RData')
```

***

# Pairwise associations - based on distance matrices

## Organize all distance matrices 

```{r}
# List of distance matrices
dist_names <- c("bc_dap_dog_id", 
                "wuf_dap_dog_id",
                "kegg_pwy_dist",
                "age_dist", 
                "weight_dist", 
                "sex_dist", 
                "genetic_distance_matrix", 
                "lab_test_dist", 
                "environment_dist", 
                "pa_dist", 
                "diet_dist",
                "clinical_dist",
                "sample_qc_dist"
                ) 

# Initialize grid for pairwise comparisons between distance matrices
dist_list <- data.frame(dist = dist_names, num = 1:length(dist_names))
comparison_grid <- cross_join(
  x = dist_list %>% rename_with(function(x) paste0(x, '_x')), 
  y = dist_list %>% rename_with(function(x) paste0(x, '_y'))
  ) %>%
  filter(num_x < num_y)

# List of distance matrices
dist_list <- lapply(dist_names, get)
names(dist_list) <- dist_names

# Pretty names for plots
pretty_names <- c("bc_dap_dog_id" = "Microbiome - Tax'", 
                  "wuf_dap_dog_id" = "Microbiome - WUF",
                  "kegg_pwy_dist" = "Microbiome - Path'",
                  "age_dist" = "Age", 
                  "weight_dist" = "Weight", 
                  "sex_dist" = "Sex", 
                  "genetic_distance_matrix" = "Genotype", 
                  "lab_test_dist" = "Blood tests",
                  "environment_dist" = "Environment", 
                  "pa_dist" = "Lifestyle",
                  "diet_dist" = "Diet",
                  "clinical_dist" = "Health conditions",
                  "sample_qc_dist" = "Seq' quality")
```

```{r}
n_perm <- 999
pairwise_mantels_backup <- 'omic_associations_backup/pairwise_mantels.csv'
pairwise_proc_backup <- 'omic_associations_backup/pairwise_protest.csv'

# Set progress bar for foreach loops
if (RECALCULATE_TESTS) {
  pb <- txtProgressBar(max = nrow(comparison_grid), style = 3)
  progress <- function(n) setTxtProgressBar(pb, n)
  opts <- list(progress = progress)
}
```

***

## Mantel test

We compare every pair of distance matrices using Mantel tests with 999 permutations. 
The final effect size is given by Spearman's rho (^2).

```{r}
if (! RECALCULATE_TESTS) {
  # Read from file
  pairwise_mantels <- read_csv(pairwise_mantels_backup, show_col_types = F)
} else {
  # Calculate Mantel between pairs of distance matrices
  pairwise_mantels <- foreach(i = 1:nrow(comparison_grid), 
                    .combine = rbind, 
                    .export = dist_names,
                    .packages = c('dplyr','stringr','ggplot2'),
                    .options.snow = opts) %dopar% {
                      
    # Get the two distances matrices to compare
    .x = comparison_grid$dist_x[i]
    .y = comparison_grid$dist_y[i]
    message("Calculating mantel for ", .x, " <> ", .y)
    dist_1 <- get(.x)
    dist_2 <- get(.y)
    
    # Run mantel
    mantel_result_spearman <- run_mantel_test(
        dist_1, 
        dist_2, 
        sample_id_column = "dap_dog_id", 
        corr_method = "spearman",
        n_permutations = n_perm
        )
    
    # Record result
    mantel_result_spearman$results %>%
      select(-n_shared_samples) %>%
      rename(mantel_corr_spear = mantel_corr, p_value_spear = p_value) %>%
      mutate(dist_x = .x, dist_y = .y) %>%
      relocate(dist_x, dist_y)
  }
  
  # Add some fields 
  pairwise_mantels <- pairwise_mantels %>%
    mutate(n_permutations = n_perm,
           var_exp_spear = pmax(0, 100*(mantel_corr_spear^2)),
           fdr_spear = p.adjust(p_value_spear, method = "fdr"))
  
  # Save 
  write_csv(pairwise_mantels, 
            file = pairwise_mantels_backup)
}
```

```{r MANTEL_PLOT, fig.height=6.2, fig.width=7, message=F}
# Organize data for plot
# Flip X and Y to have both comparison directions
tmp <- pairwise_mantels %>%
  mutate(tmp = dist_x,
         dist_x = dist_y,
         dist_y = tmp) %>%
  select(-tmp)

mantel_for_plot <- bind_rows(pairwise_mantels, tmp) %>%
  filter(dist_x %in% dist_names) %>%
  filter(dist_y %in% dist_names) %>%
  mutate(sign = get_signif_marks(fdr_spear),
         dist_x_pretty = pretty_names[dist_x],
         dist_y_pretty = pretty_names[dist_y]) %>%
  rename(var_explained = var_exp_spear)

# BC + KEGG
p_mantel_bc <- plot_pairwise_shared_var(
  mantel_for_plot %>% 
    filter((! dist_x %in% c("wuf_dap_dog_id")) & 
             (! dist_y %in% c("wuf_dap_dog_id"))),
  title = 'Pairwise associations based on\nMantel tests',
  effect_size_title = bquote('Mantel'~ rho^2~'(%)')
)
```

***

## Procrustes

```{r}
if (! RECALCULATE_TESTS) {
  # Read from file
  pairwise_protest <- read_csv(pairwise_proc_backup, show_col_types = F)
} else {
  pairwise_protest <- foreach(
    i = 1:nrow(comparison_grid), 
    .combine = rbind, 
    .export = dist_names,
    .packages = c('dplyr','ggplot2','tibble'),
    .options.snow = opts) %dopar% {
    
    # Get the two distances matrices to compare
    .x = comparison_grid$dist_x[i]
    .y = comparison_grid$dist_y[i]
    message("Calculating procrustes for ", .x, " <> ", .y)
    dist_1 <- get(.x)
    dist_2 <- get(.y)
    
    proc_result <- run_procrustes_test(
      dist_1, 
      dist_2, 
      sample_id_column = "dap_dog_id", 
      n_permutations = n_perm
    )
    
    # Return results
    proc_result$results %>% 
      mutate(dist_x = .x, dist_y = .y) %>%
      relocate(dist_x, dist_y)
  }
  
  # Add some fields 
  pairwise_protest <- pairwise_protest %>%
    mutate(n_permutations = n_perm) %>%
    mutate(fdr_p_value = p.adjust(p_value, method = "fdr")) 
  
  # Save 
  write_csv(pairwise_protest, file = pairwise_proc_backup)
}
```

```{r PROCRUSTES_PLOT, fig.height=6.2, fig.width=7}
# Organize data for plot 
tmp <- pairwise_protest %>%
  mutate(tmp = dist_x,
         dist_x = dist_y,
         dist_y = tmp) %>%
  select(-tmp)

protest_for_plot <- bind_rows(pairwise_protest, tmp) %>%
  filter(dist_x %in% dist_names) %>%
  filter(dist_y %in% dist_names) %>%
  mutate(sign = get_signif_marks(fdr_p_value),
         dist_x_pretty = pretty_names[dist_x],
         dist_y_pretty = pretty_names[dist_y]) %>%
  mutate(var_explained = 100*var_explained)

# BC
p_protest_bc <- plot_pairwise_shared_var(
  protest_for_plot %>% 
    filter((! dist_x %in% c("wuf_dap_dog_id")) & 
             (! dist_y %in% c("wuf_dap_dog_id"))),
  title = 'Pairwise associations based on\nProcrustes tests',
  effect_size_title = bquote('Procrustes'~r^2~'(%)')
)

# WUF
p_protest_wuf <- plot_pairwise_shared_var(
  protest_for_plot %>% 
    filter((! dist_x %in% c("bc_dap_dog_id")) & 
             (! dist_y %in% c("bc_dap_dog_id"))),
  title = 'Pairwise associations based on\nProcrustes tests',
  effect_size_title = bquote('Procrustes'~r^2~'(%)')
)
```

## Main figure

```{r fig.height=6, fig.width=6.2}
p_mantel_bc
ggsave(file="Figure2/Figure2C.svg", plot=p_mantel_bc, width=5.2, height=5.2)
```

## Supplementary figure

```{r fig.height=6, fig.width=6.2}
p_protest_bc
ggsave(file="SuppFigure2/SuppFigure2A.svg", plot=p_protest_bc, width=5.2, height=5.2)
```

***

# Multiple regression on distance matrices (MRM)

## Measure omic importance by MRM permutation test p value

The MRM method computes a p-value for each distance matrix included in the model, equivalent to a coefficient p-values in a linear model. Due to MRM's sensitivity to the ordering of the predictors, we evaluate these p-values across 100 random orderings.

Note: Randomness is further introduced by the permutations within MRM (so not only ordering).

```{r MRM_APPROACH2, message = FALSE}
mrm2_backup <- 'omic_associations_backup/mrm_2_backup.RData'

if (! RECALCULATE_TESTS) {
  # Read from file
  load(mrm2_backup)
} else {

  # BC
  bc_mrm2_results <- run_multiple_MRM(
    dist_list, 
    response_matrix_name = "bc_dap_dog_id",
    matrices_to_remove = c('wuf_dap_dog_id','kegg_pwy_dist'),
    title = 'Coefficient p-value of each category in a MRM\n(BC)',
    pretty_names
    )
  message('Completed MRM computation for BC distances')
  
  # WUF
  wuf_mrm2_results <- run_multiple_MRM(
    dist_list, 
    response_matrix_name = "wuf_dap_dog_id",
    matrices_to_remove = c('bc_dap_dog_id','kegg_pwy_dist'),
    title = 'Coefficient p-value of each category in a MRM\n(WUF)',
    pretty_names
    )
  message('Completed MRM computation for WUF distances')
  
  # KEGG
  kegg_mrm2_results <- run_multiple_MRM(
    dist_list,
    response_matrix_name = "kegg_pwy_dist",
    matrices_to_remove = c('wuf_dap_dog_id','bc_dap_dog_id'),
    title = 'Coefficient p-value of each category in a MRM\n(KEGG pathways)',
    pretty_names
    )
  message('Completed MRM computation for KEGG-PWY distances')
  
  save(bc_mrm2_results, 
       wuf_mrm2_results, 
       kegg_mrm2_results, 
       file = mrm2_backup)
}
```

```{r fig.width=5, fig.height=4}
bc_mrm2_results$p
wuf_mrm2_results$p
kegg_mrm2_results$p
ggsave(file="SuppFigure2/SuppFigure2B.svg", plot=bc_mrm2_results$p + ggtitle(NULL), width=5, height=3.3)
```
